---
title: "Intake Turbidity Forecast — Hierarchical GAM Proof of Concept"
author: "ROSSyndicate"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

library(here)
library(cdssr)
library(snotelr)
library(mgcv)
library(Metrics)
library(RcppRoll)
library(tidyverse)
```

# Purpose

Proof-of-concept for predicting **next-day (t+1) p95 turbidity at the FC intake**
using a hierarchical GAM structure. Given the low sample size and the two dominant
turbidity regimes — spring snowmelt and convective storms — a single global model
is likely to underfit high-turbidity events. Instead, we classify each day into one
of three regimes and fit a lightweight GAM per regime:

| Regime | Drivers | GAM predictors |
|---|---|---|
| **Snowmelt** | DOY 100–160, active melt | DOY, log(Q), melt rate |
| **Convective** | Precip event (June - ) | log(Q), ΔQ₂₄ₕ, precip features |
| **Baseflow** | Everything else | DOY, log(Q) |

All three are compared against a **global GAM** and a **persistence baseline**
(today's p95 as tomorrow's forecast) via 5-fold year-grouped cross-validation on
2010–2023. 2024–2025 is held out entirely as the test set.

All GAMs use `family = Gamma(link = "log")`, which is appropriate for strictly
positive, right-skewed turbidity data and does not require a manual log transform.

---

# Setup

```{r constants}
# ── Regime classification thresholds ─────────────────────────────────────────
PRECIP_THRESH <- 0.2   # inches: binary event indicator
CONVECTIVE_PRECIP_THRESH <- 0.5  # inches over 3 days: antecedent wetness
MELT_THRESH <- 0     # melt_rate_7d > this → active snowmelt

# ── Evaluation ────────────────────────────────────────────────────────────────
HIGH_TURB_THRESH       <- 75    # NTU: threshold for high-turbidity event evaluation

# ── Train / test split ────────────────────────────────────────────────────────
TEST_YEARS     <- 2024:2025
TRAIN_CV_YEARS <- 2010:2023

# 5 grouped-year CV folds: each fold's years are held out as validation;
# all other TRAIN_CV_YEARS are used as training. Adjust groupings as needed.
cv_folds <- list(
  fold_1 = 2010:2012,
  fold_2 = 2013:2015,
  fold_3 = 2016:2018,
  fold_4 = 2019:2020,
  fold_5 = 2021:2023
)

# ── GAM formulas ──────────────────────────────────────────────────────────────
# Global model: all predictors, all data
GLOBAL_FORMULA <- target_p95 ~
  s(doy, bs = "cc", k = 10) +
  s(deltaQ_24h, k = 10) +
  precip_event +
  s(melt_rate_7d, bs = "ts", k = 8)

# GLOBAL_FORMULA <- target_p95 ~
#   s(doy, bs = "cc", k = 10) +
#   s(log_q, k = 10) +
#   s(deltaQ_24h, k = 10) +
#   precip_3day +
#   precip_event +
#   s(melt_rate_7d, bs = "ts", k = 8)

# GLOBAL_FORMULA <- target_p95 ~
#   s(doy, bs = "cc", k = 8) +
#   s(log_q, k = 5) +
#   s(deltaQ_24h, k = 5) +
#   precip_3day +
#   precip_event +
#   s(melt_rate_7d, bs = "ts", k = 5)

# Per-regime formulas — kept simple intentionally given low regime-specific N.
# bs = "tp" for regime-specific DOY terms (data may not span full year,
# so the cyclic constraint of "cc" is inappropriate).
REGIME_FORMULAS <- list(
  snowmelt = target_p95 ~ 
    s(log_q, k = 10) +
    s(melt_rate_7d, bs = "ts", k = 8),
  convective = target_p95 ~ s(log_q, k = 8) +
    s(deltaQ_24h, k = 8) +
    precip_event,
  baseflow   = target_p95 ~ 
    s(doy, bs = "cc", k = 10) +
    s(log_q, k = 8) 
  
)
# REGIME_FORMULAS <- list(
#   snowmelt   = target_p95 ~ s(doy, bs = "tp", k = 10) +
#     s(log_q, k = 10) +
#     s(melt_rate_7d, bs = "ts", k = 8),
#   convective = target_p95 ~ s(log_q, k = 8) +
#     s(deltaQ_24h, k = 8) +
#     precip_3day +
#     precip_event,
#   baseflow   = target_p95 ~ s(doy, bs = "cc", k = 10) +
#     s(log_q, k = 8) +
#     s(deltaQ_24h, k = 8)
#   
# )

# REGIME_FORMULAS <- list(
#   snowmelt   = target_p95 ~ s(doy, bs = "tp", k = 6) +
#                              s(log_q, k = 5) +
#                              s(melt_rate_7d, bs = "ts", k = 5),
#   convective = target_p95 ~ s(log_q, k = 5) +
#                              s(deltaQ_24h, k = 5) +
#                              precip_3day +
#                              precip_event,
#   baseflow   = target_p95 ~ s(doy, bs = "cc", k = 8) +
#                              s(log_q, k = 5)
# )
```

---

# Data

```{r load-turb}
turb <- read_csv(
  here("data/collated/FC_intake/fc_intake_munroe_turbidity_quantile_record_2010-2025.csv")
) %>%
  filter(location == "intake_turb_NTU") %>%
  select(date, doy, year, p95)
```

```{r load-flow}
flow_data <- get_sw_ts(
  abbrev     = "CLAFTCCO",
  start_date = "2010-01-01",
  end_date   = "2025-12-31",
  timescale  = "day"
) %>%
  select(datetime, canyon_mouth_cfs = value) %>%
  mutate(date = as.Date(datetime))
```

```{r load-precip}
precip_data <- get_climate_ts(
  site_id    = "US1COLR0815",
  param      = "Precip",
  start_date = "2010-01-01",
  end_date   = "2025-12-31",
  timescale  = "day"
) %>%
  select(datetime, redfeather_precip = value) %>%
  mutate(date = as.Date(datetime))
```

```{r load-snotel}
# SNOTEL sites in the Cache la Poudre HUC (10190007)
snotel_sites <- snotel_info() %>%
  filter(grepl("10190007", description))

snotel_data <- snotel_download(
  site_id  = snotel_sites$site_id,
  internal = TRUE,
  metric   = TRUE
) %>%
  mutate(date = ymd(date),
         doy  = yday(date),
         year = year(date))

# PCA on SWE across sites — PC1 captures the dominant snowpack signal
snotel_matrix <- snotel_data %>%
  select(date, site_id, snow_water_equivalent) %>%
  pivot_wider(names_from  = site_id,
              names_prefix = "swe_",
              values_from  = snow_water_equivalent) %>%
  .[complete.cases(.), ]

pca_snotel <- snotel_matrix %>%
  select(starts_with("swe_")) %>%
  prcomp(center = TRUE, scale. = TRUE)

summary(pca_snotel)
pca_snotel$rotation
```

```{r snotel-melt}
# Compute SWE PC1 and melt rates (positive = losing snowpack)
snotel <- tibble(swe_pca = pca_snotel$x[, 1], date = snotel_matrix$date) %>%
  arrange(date) %>%
  mutate(date_m7  = date - days(7),
         date_m14 = date - days(14)) %>%
  left_join(
    select(., date, swe_pca) %>% rename(date_m7 = date, swe_pca_lag7 = swe_pca),
    by = "date_m7"
  ) %>%
  left_join(
    select(., date, swe_pca) %>% rename(date_m14 = date, swe_pca_lag14 = swe_pca),
    by = "date_m14"
  ) %>%
  mutate(
    melt_rate_7d  = (swe_pca_lag7  - swe_pca) / 7,
    melt_rate_14d = (swe_pca_lag14 - swe_pca) / 14
  ) %>%
  select(date, swe_pca, melt_rate_7d, melt_rate_14d)

# Confirm PC1 loads positively on all SWE sites before relying on melt_rate_7d sign
pca_snotel$rotation[, 1]  # all should be positive (or all negative — flip MELT_THRESH sign if so)

```
The PC1 sign is confirmed as expected — positive loadings!
accumulation = negative melt rate, ablation = positive.

---

# Feature Engineering & Regime Classification

```{r features}
model_data <- turb %>%
  left_join(flow_data   %>% select(date, canyon_mouth_cfs), by = "date") %>%
  left_join(precip_data %>% select(date, redfeather_precip), by = "date") %>%
  left_join(snotel,                                          by = "date") %>%
  arrange(date) %>%
  mutate(
    # Flow features
    log_q      = log1p(canyon_mouth_cfs),
    deltaQ_24h = canyon_mouth_cfs - lag(canyon_mouth_cfs, 1),
    
    # Precipitation features
    precip       = replace_na(redfeather_precip, 0),
    precip_3day  = roll_sum(replace_na(redfeather_precip, 0),
                            n = 3, fill = NA, align = "right"),
    precip_event = as.numeric(replace_na(redfeather_precip, 0) > PRECIP_THRESH),
    
    # Target: tomorrow's p95
    target_p95 = lead(p95, 1)
  ) %>%
  mutate(
    # Regime classification (priority: convective > snowmelt > baseflow)
    regime = case_when(
      # june through september
      doy %in% 150:300 & (precip_event == 1 | precip_3day > CONVECTIVE_PRECIP_THRESH)  ~ "convective",
      # jan through june
      !is.na(melt_rate_7d) & melt_rate_7d > MELT_THRESH & doy %in% 0:210 ~ "snowmelt",
      .default =  "baseflow"
    ),
    regime = factor(regime, levels = c("snowmelt", "convective", "baseflow"))
  )

# Columns required to fit/predict any of the GAM formulas
MODEL_COLS <- c("target_p95", "p95", "log_q", "deltaQ_24h",
                "precip_3day", "precip_event", "melt_rate_7d", "doy", "regime")

cat("Complete rows (all model columns):",
    sum(complete.cases(model_data[, MODEL_COLS])), "of", nrow(model_data), "\n")

model_data %>%
  filter(complete.cases(.[, MODEL_COLS])) %>%
  count(regime) %>%
  mutate(pct = round(100 * n / sum(n), 1))
```

And look at snotel data distribution

```{r}

# Distribution of melt_rate_7d by month to inform threshold choice
model_data %>%
  mutate(month = month(date)) %>%
  filter(!is.na(melt_rate_7d)) %>%
  group_by(month) %>%
  summarise(median_melt = median(melt_rate_7d), 
            q75_melt = quantile(melt_rate_7d, 0.75))
```
accumulation = negative melt rate, ablation = positive.

---

# Regime Visualisation

```{r regime-ts, fig.dim = c(10, 4)}
model_data %>%
  filter(complete.cases(.[, MODEL_COLS]), year %in% TRAIN_CV_YEARS) %>%
  ggplot(aes(x = date, y = p95, color = regime)) +
  geom_point(size = 0.5, alpha = 0.7) +
  scale_color_manual(values = c(snowmelt = "#2166ac", convective = "#d73027",
                                baseflow = "#4dac26")) +
  scale_y_continuous(limits = c(0, NA)) +
  theme_bw() +
  labs(title = "Observed p95 turbidity coloured by regime (training years)",
       x = NULL, y = "p95 turbidity (NTU)")
```

```{r regime-boxplot, fig.dim = c(6, 4)}
model_data %>%
  filter(complete.cases(.[, MODEL_COLS]), year %in% TRAIN_CV_YEARS) %>%
  ggplot(aes(x = regime, y = p95, fill = regime)) +
  geom_boxplot(outlier.size = 0.5) +
  scale_fill_manual(values = c(snowmelt = "#2166ac", convective = "#d73027",
                               baseflow = "#4dac26")) +
  scale_y_log10() +
  theme_bw() +
  theme(legend.position = "none") +
  labs(title = "p95 turbidity by regime (log scale)", x = NULL, y = "p95 NTU (log)")
```

---

# Cross-Validation

```{r cv-function}
# Fits persistence, global GAM, and all three regime GAMs for one fold.
# Returns a list with:
#   $predictions  — tidy tibble of per-observation predictions on the val set
#   $models       — list(global = ..., regime = list(...)) of fitted GAMs
#   $train_data   — training data frame (used later for response curve medians)
run_cv_fold <- function(k) {
  val_yrs   <- cv_folds[[k]]
  train_yrs <- setdiff(TRAIN_CV_YEARS, val_yrs)
  
  train <- model_data %>%
    filter(year %in% train_yrs) %>%
    drop_na(all_of(MODEL_COLS))
  
  val <- model_data %>%
    filter(year %in% val_yrs) %>%
    drop_na(all_of(MODEL_COLS))
  
  if (nrow(val) == 0) return(NULL)
  
  # ── Global GAM ───────────────────────────────────────────────────────────────
  fit_global  <- gam(GLOBAL_FORMULA, data = train,
                     family = Gamma(link = "log"), method = "REML")
  global_pred <- as.numeric(predict(fit_global, newdata = val, type = "response"))
  
  # ── Regime GAMs — fit all regimes on training data ───────────────────────────
  # Fit every regime regardless of whether it appears in val (needed for test
  # ensemble later). Fall back to global GAM if too few training rows or error.
  regime_fits <- map(levels(train$regime), function(r) {
    train_r <- filter(train, regime == r)
    if (nrow(train_r) < 10) return(fit_global)
    tryCatch(
      gam(REGIME_FORMULAS[[r]], data = train_r,
          family = Gamma(link = "log"), method = "REML"),
      error = function(e) fit_global
    )
  }) %>% set_names(levels(train$regime))
  
  # ── Hierarchical val predictions ─────────────────────────────────────────────
  hier_pred <- rep(NA_real_, nrow(val))
  for (r in levels(val$regime)) {
    idx_r <- which(val$regime == r)
    if (length(idx_r) == 0) next
    hier_pred[idx_r] <- as.numeric(
      predict(regime_fits[[r]], newdata = val[idx_r, ], type = "response")
    )
  }
  
  list(
    predictions = tibble(
      fold             = k,
      date             = val$date,
      regime           = val$regime,
      observed         = val$target_p95,
      persistence      = val$p95,
      global_gam       = global_pred,
      hierarchical_gam = hier_pred
    ),
    models     = list(global = fit_global, regime = regime_fits),
    train_data = train
  )
}
```

```{r run-cv}
cv_out     <- map(seq_along(cv_folds), run_cv_fold)
cv_results <- map(cv_out, "predictions") %>% bind_rows()
```

## CV Metrics Summary

```{r cv-metrics}
cv_long <- cv_results %>%
  pivot_longer(cols = c(persistence, global_gam, hierarchical_gam),
               names_to = "model", values_to = "predicted") %>%
  filter(!is.na(predicted))

# Per-fold metrics
fold_metrics <- cv_long %>%
  group_by(model, fold) %>%
  summarise(
    rmse = rmse(observed, predicted),
    mae  = mae(observed, predicted),
    bias = mean(predicted - observed),
    .groups = "drop"
  )

fold_metrics %>%
  group_by(model) %>%
  summarise(
    rmse_mean = mean(rmse), rmse_sd = sd(rmse),
    mae_mean  = mean(mae),  mae_sd  = sd(mae),
    bias_mean = mean(bias),
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2))) %>%
  arrange(rmse_mean)
```

```{r cv-scatter, fig.dim = c(9, 3)}
cv_long %>%
  mutate(model = recode(model,
                        persistence      = "Persistence",
                        global_gam       = "Global GAM",
                        hierarchical_gam = "Hierarchical GAM"
  )) %>%
  ggplot(aes(x = observed, y = predicted, color = regime)) +
  geom_point(alpha = 0.3, size = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  scale_color_manual(values = c(snowmelt = "#2166ac", convective = "#d73027",
                                baseflow = "#4dac26")) +
  facet_wrap(~ model) +
  theme_bw() +
  labs(title = "CV: Predicted vs Observed p95 turbidity (NTU)",
       x = "Observed (NTU)", y = "Predicted (NTU)")
```

```{r cv-highturb}
# Performance restricted to high-turbidity events
cv_long %>%
  filter(observed >= HIGH_TURB_THRESH) %>%
  group_by(model) %>%
  summarise(
    n    = n(),
    rmse = rmse(observed, predicted),
    mae  = mae(observed, predicted),
    bias = mean(predicted - observed),
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2))) %>%
  arrange(rmse) %>%
  print()
```

```{r cv-by-regime}
# Per-regime breakdown for the hierarchical model
cv_long %>%
  filter(model == "hierarchical_gam") %>%
  group_by(regime) %>%
  summarise(
    n    = n(),
    rmse = rmse(observed, predicted),
    mae  = mae(observed, predicted),
    bias = mean(predicted - observed),
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))
```

## CV Validation Time Series

One panel per fold, showing observed p95 (coloured by regime), hierarchical GAM
prediction (black line), and persistence (dashed). Useful for spotting whether
the model tracks event timing and magnitude across different years.

```{r cv-ts, fig.dim = c(10, 4)}
cv_results %>%
  mutate(fold_label = paste0("Fold ", fold, " (", map_chr(fold, ~ paste(cv_folds[[.x]], collapse = "–")), ")")) %>%
  ggplot(aes(x = date)) +
  geom_point(aes(y = observed, color = regime), size = 0.8, alpha = 0.8) +
  geom_line(aes(y = hierarchical_gam), color = "black", linewidth = 0.5) +
  geom_line(aes(y = global_gam), color = "orange", linewidth = 0.5) +
  #geom_line(aes(y = persistence), color = "grey50", linewidth = 0.4, linetype = "dashed") +
  geom_hline(yintercept = HIGH_TURB_THRESH, linetype = "dotted", color = "grey40") +
  scale_color_manual(values = c(snowmelt = "#2166ac", convective = "#d73027",
                                baseflow = "#4dac26")) +
  scale_y_continuous(limits = c(0, NA)) +
  facet_wrap(~ fold_label, scales = "free_x", ncol = 1) +
  coord_cartesian(ylim = c(0, 250)) +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(
    title = "CV validation folds: observed (coloured), hierarchical GAM (black), persistence (dashed)",
    x = NULL, y = "p95 turbidity (NTU)"
  )
```

---

# CV Model Inspection

Rather than fitting a separate final model on all training data, we use the
ensemble of CV fold models throughout — for response curves, diagnostics, and
test set prediction. This incorporates interannual variability across the full
training record and avoids a separate train/inspect split.

## Response Curves

Each line is one CV fold's fitted response; the spread across folds reflects
how much the estimated effect varies with different training years excluded.
Other predictors are held at the fold's regime-specific training median.

```{r gam-response-helper}
# Predict along one predictor, others at training median.
# Returns predictions on response scale with 95% CI.
pred_curve <- function(model, data, var, n = 80) {
  base <- data %>%
    summarise(across(where(is.numeric), ~median(.x, na.rm = TRUE)))
  grid <- base %>%
    slice(rep(1, n)) %>%
    mutate(!!var := seq(min(data[[var]], na.rm = TRUE),
                        max(data[[var]], na.rm = TRUE),
                        length.out = n))
  p <- predict(model, newdata = grid, type = "link", se.fit = TRUE)
  grid %>%
    mutate(predicted = exp(p$fit),
           lower     = exp(p$fit - 1.96 * p$se.fit),
           upper     = exp(p$fit + 1.96 * p$se.fit))
}
```

```{r gam-flow-response, fig.dim = c(9, 3)}
# Flow response per regime — all CV folds overlaid.
flow_curves <- map_dfr(seq_along(cv_out), function(k) {
  imap_dfr(cv_out[[k]]$models$regime, function(fit, nm) {
    train_r <- filter(cv_out[[k]]$train_data, regime == nm)
    if (nrow(train_r) < 10) return(NULL)
    pred_curve(fit, train_r, "log_q") %>%
      mutate(regime = nm, fold = k, cfs = expm1(log_q))
  })
})

ggplot(flow_curves, aes(x = cfs, y = predicted,
                        color = regime, group = interaction(fold, regime))) +
  geom_line(alpha = 0.4, linewidth = 0.6) +
  scale_color_manual(values = c(snowmelt = "#2166ac", convective = "#d73027",
                                baseflow = "#4dac26")) +
  scale_x_log10(labels = scales::comma) +
  facet_wrap(~ regime, scales = "free") +
  theme_bw() +
  theme(legend.position = "none") +
  labs(title = "Regime GAMs: fitted turbidity vs flow — all CV folds",
       x = "Canyon mouth flow (CFS, log scale)", y = "Predicted p95 (NTU)")
```

```{r gam-doy-response, fig.dim = c(6, 3)}
# Seasonal baseline — global GAM DOY effect, all CV folds.
doy_curves <- map_dfr(seq_along(cv_out), function(k) {
  pred_curve(cv_out[[k]]$models$global, cv_out[[k]]$train_data, "doy") %>%
    mutate(fold = k)
})

ggplot(doy_curves, aes(x = doy, y = predicted, group = fold)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.08, fill = "grey40") +
  geom_line(alpha = 0.5, linewidth = 0.6, color = "grey30") +
  theme_bw() +
  labs(title = "Global GAM: seasonal baseline — all CV folds",
       x = "Day of year", y = "Predicted p95 (NTU)")
```

```{r gam-melt-response, fig.dim = c(7, 3)}
# Snowmelt GAM: melt rate effect at median flow — all CV folds.
melt_curves <- map_dfr(seq_along(cv_out), function(k) {
  train_r <- filter(cv_out[[k]]$train_data, regime == "snowmelt")
  if (nrow(train_r) < 10) return(NULL)
  pred_curve(cv_out[[k]]$models$regime$snowmelt, train_r, "melt_rate_7d") %>%
    mutate(fold = k)
})

ggplot(melt_curves, aes(x = melt_rate_7d, y = predicted, group = fold)) +
  geom_line(alpha = 0.5, linewidth = 0.6, color = "#2166ac") +
  theme_bw() +
  labs(title = "Snowmelt GAM: turbidity vs 7-day melt rate (median flow) — all CV folds",
       x = "7-day melt rate (SWE PC1 units / day)", y = "Predicted p95 (NTU)")
```

```{r gam-convective-precip, fig.dim = c(7, 3)}
# Convective GAM: antecedent precip effect — event vs non-event, all CV folds.
precip_curves <- map_dfr(seq_along(cv_out), function(k) {
  train_r <- filter(cv_out[[k]]$train_data, regime == "convective")
  if (nrow(train_r) < 10) return(NULL)
  map_dfr(c(0, 1), function(evt) {
    base <- train_r %>%
      summarise(across(where(is.numeric), ~median(.x, na.rm = TRUE))) %>%
      mutate(precip_event = evt)
    grid <- base %>%
      slice(rep(1, 80)) %>%
      mutate(precip_3day = seq(0, quantile(train_r$precip_3day, 0.95, na.rm = TRUE),
                               length.out = 80))
    p <- predict(cv_out[[k]]$models$regime$convective,
                 newdata = grid, type = "link", se.fit = TRUE)
    grid %>%
      mutate(predicted   = exp(p$fit),
             fold        = k,
             event_label = if_else(evt == 1, "Precip event day", "No event"))
  })
})

ggplot(precip_curves,
       aes(x = precip_3day, y = predicted,
           color = event_label, group = interaction(fold, event_label))) +
  geom_line(alpha = 0.45, linewidth = 0.6) +
  scale_color_manual(values = c("Precip event day" = "#d73027", "No event" = "grey40"),
                     name = NULL) +
  theme_bw() +
  labs(title = "Convective GAM: turbidity vs antecedent precip — all CV folds",
       x = "3-day rolling precip (inches)", y = "Predicted p95 (NTU)")
```

## Model Diagnostics

Compact edf/k' table across all folds to flag any smooths being constrained
(edf ≈ k' − 1 suggests `k` should be increased). Full diagnostics shown for
fold 3 as a representative example.

```{r diag-edf-table}
# edf and k' for every smooth term, every fold and regime
map_dfr(seq_along(cv_out), function(k) {
  # Regime GAMs
  imap_dfr(cv_out[[k]]$models$regime, function(fit, nm) {
    st <- summary(fit)$s.table
    tibble(fold = k, model = paste0("regime:", nm),
           term = rownames(st), edf = st[, "edf"], k_prime = st[, "Ref.df"])
  }) %>%
    bind_rows(
      # Global GAM
      {
        st <- summary(cv_out[[k]]$models$global)$s.table
        tibble(fold = k, model = "global",
               term = rownames(st), edf = st[, "edf"], k_prime = st[, "Ref.df"])
      }
    )
}) %>%
  mutate(constrained = edf > k_prime - 1) %>%
  group_by(model, term) %>%
  summarise(edf_mean  = round(mean(edf), 2),
            edf_range = paste0(round(min(edf), 1), "–", round(max(edf), 1)),
            k_prime   = round(mean(k_prime), 1),
            any_constrained = any(constrained),
            .groups = "drop") %>%
  arrange(model, term) %>%
  print(n = Inf)
```

```{r diag-fold3, fig.dim = c(9, 4)}
# Full gam.check and partial effects for fold 3 (representative)
cat("── Global GAM, fold 3 ──\n")
summary(cv_out[[3]]$models$global)
gam.check(cv_out[[3]]$models$global)
plot(cv_out[[3]]$models$global, pages = 1, residuals = TRUE, shade = TRUE,
     main = "Global GAM partial effects, fold 3 (link scale)")

walk2(cv_out[[3]]$models$regime, names(cv_out[[3]]$models$regime), function(fit, nm) {
  n_r <- filter(cv_out[[3]]$train_data, regime == nm) %>% nrow()
  cat("\n── Regime:", nm, "| fold 3 training n =", n_r, "──\n")
  print(summary(fit))
  plot(fit, pages = 1, residuals = TRUE, shade = TRUE,
       main = paste("Partial effects —", nm, ", fold 3 (link scale)"))
})
```

---

# Test Set Evaluation (2024–2025)

Predictions are the **mean across all 5 CV fold models** (ensemble). The
fold-to-fold SD is retained as a rough indicator of prediction uncertainty
from interannual variability in the training data.

```{r test-eval}
test_df <- model_data %>%
  filter(year %in% TEST_YEARS) %>%
  drop_na(all_of(MODEL_COLS))

# Predict using each fold's models
test_by_fold <- map_dfr(seq_along(cv_out), function(k) {
  fold_models <- cv_out[[k]]$models
  
  global_pred <- as.numeric(
    predict(fold_models$global, newdata = test_df, type = "response")
  )
  
  hier_pred <- rep(NA_real_, nrow(test_df))
  for (r in levels(test_df$regime)) {
    idx_r <- which(test_df$regime == r)
    if (length(idx_r) == 0) next
    hier_pred[idx_r] <- as.numeric(
      predict(fold_models$regime[[r]], newdata = test_df[idx_r, ], type = "response")
    )
  }
  
  tibble(fold = k, date = test_df$date, regime = test_df$regime,
         observed = test_df$target_p95, persistence = test_df$p95,
         global_gam = global_pred, hierarchical_gam = hier_pred)
})

# Ensemble: mean and SD across folds
test_results <- test_by_fold %>%
  group_by(date, regime, observed, persistence) %>%
  summarise(
    global_ensemble = mean(global_gam,       na.rm = TRUE),
    hier_ensemble   = mean(hierarchical_gam, na.rm = TRUE),
    hier_sd         = sd(hierarchical_gam,   na.rm = TRUE),
    .groups = "drop"
  )

# ── Metrics ──────────────────────────────────────────────────────────────────
test_results %>%
  pivot_longer(c(persistence, global_ensemble, hier_ensemble),
               names_to = "model", values_to = "predicted") %>%
  group_by(model) %>%
  summarise(
    n    = n(),
    rmse = rmse(observed, predicted),
    mae  = mae(observed, predicted),
    bias = mean(predicted - observed),
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2))) %>%
  arrange(rmse)
```

```{r test-scatter, fig.dim = c(9, 3)}
test_results %>%
  pivot_longer(c(persistence, global_ensemble, hier_ensemble),
               names_to = "model", values_to = "predicted") %>%
  mutate(model = recode(model,
                        persistence      = "Persistence",
                        global_ensemble  = "Global GAM ensemble",
                        hier_ensemble    = "Hierarchical GAM ensemble"
  )) %>%
  ggplot(aes(x = observed, y = predicted, color = regime)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  scale_color_manual(values = c(snowmelt = "#2166ac", convective = "#d73027",
                                baseflow = "#4dac26")) +
  facet_wrap(~ model) +
  theme_bw() +
  labs(title = "Test set (2024–2025): Predicted vs Observed p95 turbidity",
       x = "Observed (NTU)", y = "Predicted (NTU)")
```

```{r test-ts, fig.dim = c(10, 4)}
# Ribbon shows ±1 SD across fold predictions (ensemble spread)
test_results %>%
  ggplot(aes(x = date)) +
  geom_point(aes(y = observed, color = regime), size = 1, alpha = 0.8) +
  geom_ribbon(aes(ymin = hier_ensemble - hier_sd,
                  ymax = hier_ensemble + hier_sd), alpha = 0.15) +
  geom_line(aes(y = hier_ensemble), color = "black", linewidth = 0.6) +
  geom_line(aes(y = persistence), color = "grey50", linewidth = 0.4, linetype = "dashed") +
  geom_hline(yintercept = HIGH_TURB_THRESH, linetype = "dotted", color = "grey40") +
  scale_color_manual(values = c(snowmelt = "#2166ac", convective = "#d73027",
                                baseflow = "#4dac26")) +
  scale_y_continuous(limits = c(0, NA)) +
  theme_bw() +
  labs(
    title = "Test period: observed (coloured), hierarchical ensemble (black ± fold SD), persistence (dashed)",
    x = NULL, y = "p95 turbidity (NTU)"
  )
```

```{r test-highturb}
test_results %>%
  filter(observed >= HIGH_TURB_THRESH) %>%
  pivot_longer(c(persistence, global_ensemble, hier_ensemble),
               names_to = "model", values_to = "predicted") %>%
  group_by(model) %>%
  summarise(
    n    = n(),
    rmse = rmse(observed, predicted),
    mae  = mae(observed, predicted),
    bias = mean(predicted - observed),
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2))) %>%
  arrange(rmse)
```

---

# Export Models

```{r export}
dir.create(here("data/derived/models/turbidity"), showWarnings = FALSE, recursive = TRUE)

today <- format(Sys.Date(), "%Y-%m-%d")

# Save all CV fold models as a list — operational prediction averages across all
# folds (see Notes for prediction workflow)
write_rds(map(cv_out, "models"),
          here(paste0("data/derived/models/turbidity/turbidity_intake_GAM_cv_ensemble_v", today, ".rds")))

# Save the SWE PCA model — required to compute melt_rate_7d for new predictions
write_rds(pca_snotel,
          here(paste0("data/derived/models/turbidity/SWE_PCA_v", today, ".rds")))
```

---

# Notes

**Model selection:** Compare global vs hierarchical on both CV and test RMSE,
particularly on high-turbidity events (`observed >= HIGH_TURB_THRESH`). The
hierarchical model is preferred if it improves high-turbidity performance without
degrading baseflow performance, since getting high-turbidity events right matters
most operationally.

**Regime boundary tuning:** The thresholds in `constants` (`MELT_THRESH`,
`CONVECTIVE_PRECIP_THRESH`, `PRECIP_THRESH`) are initial estimates. Worth
adjusting if the regime counts (see `model_data %>% count(regime)`) produce
very unequal splits or if regime-specific model diagnostics look poor.

**GAM complexity:** All formulas use modest `k` values (5–8) given the limited
per-regime N. Run `gam.check()` on each final model — if edf ≈ k' for any smooth,
the effective degrees of freedom are being constrained and `k` should be increased.
Do this cautiously; the model should remain conservative.

**Gamma(link = "log") caution:** Predictions extrapolate on a log scale and can
grow rapidly. Check for implausibly high predictions on the test set, especially
during the convective regime. If prediction intervals are needed operationally,
use `predict(..., se.fit = TRUE)` and propagate the SE on the log scale.

**Features to consider adding:**
- Munroe turbidity as an upstream leading indicator (check data availability)
- 14-day melt rate (see TOC baseline — showed value there)
- Longer antecedent precip windows (7-day sum)

**Operational use:** To generate a forecast on day *t*:

1. Load `cv_ensemble` (list of 5 fold model lists) and `pca_snotel`
2. Observe today's canyon mouth flow, precip, and SNOTEL SWE
3. Project SNOTEL SWE onto `pca_snotel` to get `swe_pca`; compute `melt_rate_7d`
4. Classify regime using thresholds above
5. For each fold model, predict with the appropriate regime GAM; average the 5 predictions:
```r
mean(map_dbl(cv_ensemble, ~ predict(.x$regime[[regime]], newdata = today_features, type = "response")))
```
