---
title: "00_functions — Daily p95 Turbidity Forecast (Modified Kemper)"
author: "ROSSyndicate"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

Helper functions shared across all scripts in the modified Kemper daily p95
turbidity forecasting workflow. Source this file at the top of each script via
`knitr::knit("00_functions.Rmd", tangle = TRUE)` or inline with
`source(knitr::purl("00_functions.Rmd", quiet = TRUE))`.

Key differences from the original Kemper `00_functions.Rmd`:
- All functions operate on **daily** data (not hourly).
- Flow is in CFS (from CDSSR CLAFTCCO); no CMS conversion needed.
- Target is **daily p95 turbidity (NTU)**, not instantaneous FNU.
- Snow/melt features (`melt_rate_calculator`) are added for the CLP environment.
- GAM cross-validation replaces the quantile predictor functions.

---

```{r packages}
library(tidyverse)
library(slider)
library(mgcv)
library(hydroGOF)
library(lightgbm)
library(tidymodels)
library(bonsai)
```

---

# Antecedent Flow Feature Builder

```{r antecedent_Q_maker}

#### antecedent_Q_maker
####
#### Given a daily tibble with columns `date` and `canyon_mouth_cfs`,
#### computes the full suite of antecedent discharge features used as
#### predictors in both the GAM and LightGBM models.
####
#### Adapted from kemper_application/01_data_prep.Rmd antecedent calculations,
#### but simplified to daily resolution.
####
#### Returns the input tibble with additional columns appended.

antecedent_Q_maker <- function(df) {

  df %>%
    arrange(date) %>%
    mutate(
      # Log-transform (log1p guards against Q = 0)
      log_q               = log1p(canyon_mouth_cfs),
      log_q_sq            = log_q^2,

      # Day-over-day log change (hydrograph limb indicator)
      # Positive = rising, negative = falling
      log_limb            = log_q - lag(log_q, 1),

      # Normalised flow change: ΔQ / Q (Kemper eq. 8)
      dq_q                = (canyon_mouth_cfs - lag(canyon_mouth_cfs, 1)) /
                              lag(canyon_mouth_cfs, 1),

      # Raw flow change over 1 and 2 days
      deltaQ_24h          = canyon_mouth_cfs - lag(canyon_mouth_cfs, 1),
      deltaQ_48h          = canyon_mouth_cfs - lag(canyon_mouth_cfs, 2),

      # Rolling means (right-aligned, current day included)
      mean_q_7d           = slide_dbl(canyon_mouth_cfs, mean,
                                      .before = 6,  .complete = TRUE),
      mean_q_30d          = slide_dbl(canyon_mouth_cfs, mean,
                                      .before = 29, .complete = TRUE),
      mean_q_90d          = slide_dbl(canyon_mouth_cfs, mean,
                                      .before = 89, .complete = TRUE),

      # Rolling maxima
      max_q_30d           = slide_dbl(canyon_mouth_cfs, max,
                                      .before = 29, .complete = TRUE),
      max_q_90d           = slide_dbl(canyon_mouth_cfs, max,
                                      .before = 89, .complete = TRUE),
      max_q_365d          = slide_dbl(canyon_mouth_cfs, max,
                                      .before = 364, .complete = TRUE)
    ) %>%
    # Days since each rolling-maximum peak
    group_by(year = lubridate::year(date)) %>%
    mutate(
      time_since_annual_peak = as.numeric(date - date[which.max(max_q_365d)])
    ) %>%
    ungroup() %>%
    mutate(
      time_since_30d_peak = as.numeric(
        date - date[cumsum(!is.na(max_q_30d)) %>%
                      match(seq_along(date))]
      ),
      time_since_90d_peak = as.numeric(
        date - date[cumsum(!is.na(max_q_90d)) %>%
                      match(seq_along(date))]
      )
    )
}
```

---

# Snowmelt Rate Calculator

```{r melt_rate_calculator}

#### melt_rate_calculator
####
#### Given a tibble with columns `date` and `swe_pc1` (PCA score from SNOTEL
#### network), computes forward-difference melt rates at 7- and 14-day windows.
####
#### Sign convention (confirmed from PCA rotation in intake_turbidity_forecast_poc.Rmd):
####   melt_rate_7d > 0  →  active snowpack depletion (ablation)
####   melt_rate_7d < 0  →  snowpack accumulation
####
#### Returns the input tibble with `melt_rate_7d` and `melt_rate_14d` appended.

melt_rate_calculator <- function(df) {

  df %>%
    arrange(date) %>%
    mutate(
      swe_pc1_lag7  = lag(swe_pc1, 7),
      swe_pc1_lag14 = lag(swe_pc1, 14),
      melt_rate_7d  = (swe_pc1_lag7  - swe_pc1) / 7,
      melt_rate_14d = (swe_pc1_lag14 - swe_pc1) / 14
    ) %>%
    select(-swe_pc1_lag7, -swe_pc1_lag14)
}
```

---

# Regime Classifier

```{r regime_classifier}

#### regime_classifier
####
#### Classifies each row as one of three turbidity-generation regimes.
#### Classification mirrors the logic in intake_turbidity_forecast_poc.Rmd
#### but is exposed as a function for reuse in feature engineering and
#### forecast preparation.
####
#### Priority: convective > snowmelt > baseflow
####
#### Parameters (thresholds can be overridden):
####   PRECIP_THRESH              daily precip (in) for binary event flag
####   CONVECTIVE_PRECIP_THRESH   3-day rolling precip (in) for antecedent wetness
####   MELT_THRESH                melt_rate_7d threshold for active snowmelt
####
#### Required columns: doy, precip_event, precip_3day, melt_rate_7d
#### Returns the input tibble with a `regime` factor column appended.

PRECIP_THRESH            <- 0.2
CONVECTIVE_PRECIP_THRESH <- 0.5
MELT_THRESH              <- 0

regime_classifier <- function(df,
                              precip_thresh   = PRECIP_THRESH,
                              conv_precip_3d  = CONVECTIVE_PRECIP_THRESH,
                              melt_thresh     = MELT_THRESH) {
  df %>%
    mutate(
      regime = case_when(
        # Convective: June–October (DOY 150–300) with precip trigger
        doy %in% 150:300 &
          (precip_event == 1 | precip_3day > conv_precip_3d) ~ "convective",
        # Snowmelt: Jan–late July (DOY 1–210), active melt, no precip trigger
        !is.na(melt_rate_7d) &
          melt_rate_7d > melt_thresh &
          doy %in% 1:210 ~ "snowmelt",
        # Baseflow: everything else
        .default = "baseflow"
      ),
      regime = factor(regime, levels = c("snowmelt", "convective", "baseflow"))
    )
}
```

---

# GAM Cross-Validation

```{r run_gam_cv_fold}

#### run_gam_cv_fold
####
#### Fits persistence, global GAM, and regime-specific GAMs for one CV fold.
#### Adapted directly from intake_turbidity_forecast_poc.Rmd `run_cv_fold()`.
####
#### Arguments:
####   k            name of the fold in cv_folds list
####   model_data   full modelling tibble (must include MODEL_COLS)
####   cv_folds     named list of year vectors (held-out validation years per fold)
####   TRAIN_CV_YEARS  integer vector of all training+CV years
####   MODEL_COLS   character vector of required column names
####   GLOBAL_FORMULA   formula for global GAM
####   REGIME_FORMULAS  named list of formulas (snowmelt, convective, baseflow)
####   lead          integer lead time (1, 2, or 3)
####
#### Returns a tibble of per-observation predictions for the validation set.

run_gam_cv_fold <- function(k,
                            model_data,
                            cv_folds,
                            TRAIN_CV_YEARS,
                            MODEL_COLS,
                            GLOBAL_FORMULA,
                            REGIME_FORMULAS,
                            lead = 1) {

  target_col  <- paste0("target_p95_t", lead)
  val_yrs     <- cv_folds[[k]]
  train_yrs   <- setdiff(TRAIN_CV_YEARS, val_yrs)

  train <- model_data %>%
    filter(year %in% train_yrs) %>%
    drop_na(all_of(c(MODEL_COLS, target_col)))

  val <- model_data %>%
    filter(year %in% val_yrs) %>%
    drop_na(all_of(c(MODEL_COLS, target_col)))

  if (nrow(val) == 0) return(NULL)

  # Rename target to generic `target_p95` so formulas work for any lead
  train <- train %>% rename(target_p95 = !!target_col)
  val   <- val   %>% rename(target_p95 = !!target_col)

  # Global GAM
  fit_global   <- gam(GLOBAL_FORMULA, data = train,
                      family = Gamma(link = "log"), method = "REML")
  global_pred  <- as.numeric(predict(fit_global, newdata = val,
                                     type = "response"))

  # Regime GAMs — fit on all-regime training data, predict by regime
  regime_models <- map(REGIME_FORMULAS, function(f) {
    gam(f, data = train, family = Gamma(link = "log"), method = "REML")
  })

  regime_pred <- val %>%
    select(date, regime, target_p95) %>%
    mutate(
      pred = map2_dbl(
        as.character(regime), seq_len(n()),
        ~ as.numeric(predict(regime_models[[.x]],
                             newdata = val[.y, ],
                             type = "response"))
      )
    )

  tibble(
    date       = val$date,
    year       = val$year,
    lead       = lead,
    fold       = k,
    regime     = val$regime,
    observed   = val$target_p95,
    persistence = val$p95,
    global_pred = global_pred,
    regime_pred = regime_pred$pred
  )
}
```

---

# Error / Performance Metrics

```{r error_estimator}

#### error_estimator
####
#### Computes standard performance metrics for a model prediction vector.
#### Copied and extended from kemper_application/02_model_development.Rmd.

decompose_kge <- function(sim, obs) {
  r          <- cor(sim, obs, method = "pearson", use = "pairwise.complete.obs")
  bias       <- mean(sim, na.rm = TRUE) / mean(obs, na.rm = TRUE)
  variability <- sd(sim, na.rm = TRUE) / sd(obs, na.rm = TRUE)
  kge_manual <- 1 - sqrt((r - 1)^2 + (variability - 1)^2 + (bias - 1)^2)
  tibble(r, variability, bias, kge_manual)
}

error_estimator <- function(pred, obs, model_name) {
  raw_err <- pred - obs
  abs_err <- abs(raw_err)
  sqr_err <- raw_err^2
  pct_err <- abs_err / obs

  tibble(
    model  = model_name,
    rmse   = sqrt(mean(sqr_err, na.rm = TRUE)),
    mae    = mean(abs_err,  na.rm = TRUE),
    mape   = mean(pct_err,  na.rm = TRUE),
    nse    = NSE(pred, obs),
    kge    = KGE(pred, obs),
    pbias  = pbias(pred, obs),
    corr   = cor(pred, obs, use = "complete.obs")
  ) %>%
    bind_cols(decompose_kge(pred, obs))
}
```

---

# LightGBM Tuner

```{r lgbm_tuner_daily}

#### lgbm_tuner_daily
####
#### Backwards variable selection + hyperparameter tuning for daily LightGBM.
#### Adapted from kemper_application/02_model_development.Rmd.
####
#### Variable selection: expanding-window CV over 2 folds
####   WY2012–WY2020 train → WY2021 validate
####   WY2012–WY2021 train → WY2022 validate
####
#### Hyperparameter tuning: expanding-window CV over 3 folds
####   WY2012–WY2019 → WY2020, WY2012–WY2020 → WY2021, WY2012–WY2021 → WY2022
####
#### Arguments:
####   df            daily tibble with `year` + all candidate predictors + log_p95
####   all_preds     character vector of all candidate predictor column names
####   response_col  name of the log-transformed response column (default: "log_p95")
####
#### Returns a list: $best_predictors, $best_params, $varsel_stats, $tune_scores

lgbm_tuner_daily <- function(df,
                             all_preds,
                             response_col = "log_p95") {

  lgbm_predictors <- df %>%
    select(all_of(c(response_col, "year", all_preds))) %>%
    mutate(year2 = year) %>%
    filter(is.finite(.data[[response_col]]))

  lgbm_default_params <- list(
    objective    = "regression",
    num_leaves   = 31L,
    learning_rate = 0.1,
    min_data_in_leaf = 20L,
    num_threads  = 4L,
    verbose      = -1L
  )

  model_stats_all <- list()
  lgbm_var_imp    <- list()
  vars_list       <- list()

  # ── Backwards variable selection ─────────────────────────────────────────────
  for (i in seq_len(ncol(lgbm_predictors) - 3L)) {  # -3: response, year, year2

    run <- paste0("run", i)

    for (j in 2021:2022) {

      train_d <- lgbm_predictors %>%
        filter(year >= 2012, year < j) %>%
        select(!c(!!response_col, year, year2))

      test_d  <- lgbm_predictors %>%
        filter(year == j)

      preds_mat    <- data.matrix(train_d)
      response_vec <- lgbm_predictors %>%
        filter(year >= 2012, year < j) %>%
        pull(!!response_col)
      test_mat     <- data.matrix(test_d %>%
                                    select(!c(!!response_col, year, year2)))

      lgbm_ds <- lgb.Dataset(preds_mat, label = response_vec)
      set.seed(913)
      lgbm_mod <- lgb.train(lgbm_default_params, data = lgbm_ds,
                            nrounds = 100L, verbose = -1L)

      # Smearing factor
      smear_d   <- mean(10^(response_vec -
                              predict(lgbm_mod, newdata = preds_mat)),
                        na.rm = TRUE)
      log_pred  <- predict(lgbm_mod, newdata = test_mat)
      pred_turb <- (10^log_pred) * smear_d
      obs_turb  <- 10^test_d[[response_col]]

      stats_row <- error_estimator(pred_turb, obs_turb,
                                   paste0(run, "_WY", j))
      model_stats_all[[paste0(run, "_WY", j)]] <- stats_row

      imp <- lgb.importance(lgbm_mod, percentage = TRUE) %>%
        as_tibble() %>%
        mutate(run = run, valid_wy = j)
      lgbm_var_imp[[paste0(run, "_WY", j)]] <- imp
    }

    mean_imp <- bind_rows(lgbm_var_imp[grep(run, names(lgbm_var_imp))]) %>%
      group_by(Feature) %>%
      summarise(mean_gain = mean(Gain, na.rm = TRUE)) %>%
      arrange(mean_gain)

    vars_list[[run]] <- lgbm_predictors %>%
      select(!c(!!response_col, year, year2)) %>%
      colnames()

    drop_var        <- mean_imp$Feature[[1]]
    lgbm_predictors <- lgbm_predictors %>% select(-all_of(drop_var))

    if ((ncol(lgbm_predictors) - 3L) <= 1L) break
  }

  varsel_stats <- bind_rows(model_stats_all, .id = "run") %>%
    group_by(run) %>%
    summarise(mean_nse   = mean(nse,   na.rm = TRUE),
              mean_kge   = mean(kge,   na.rm = TRUE),
              mean_pbias = mean(pbias, na.rm = TRUE)) %>%
    filter(mean_nse > 0, mean_kge > 0, abs(mean_pbias) < 25) %>%
    arrange(desc(mean_kge))

  best_run        <- varsel_stats$run[[1]]
  best_predictors <- vars_list[[best_run]]

  # ── Hyperparameter tuning ─────────────────────────────────────────────────────
  lgbm_tune_data <- df %>%
    select(all_of(c(response_col, "year", best_predictors))) %>%
    mutate(year2 = year) %>%
    filter(is.finite(.data[[response_col]]))

  splits_list <- list()
  for (h in 2020:2022) {
    train_d <- lgbm_tune_data %>%
      filter(year >= 2012, year < h) %>%
      select(!c(year, year2))
    valid_d <- lgbm_tune_data %>%
      filter(year == h) %>%
      select(!c(year, year2))
    splits_list[[2023 - h]] <- make_splits(train_d, assessment = valid_d)
  }
  tune_resamples <- manual_rset(splits_list, c("Split1", "Split2", "Split3"))

  tune_recipe <- recipe(
    as.formula(paste(response_col, "~ .")),
    data = lgbm_tune_data %>%
      filter(year >= 2012, year < 2020) %>%
      select(!c(year, year2))
  ) %>% prep()

  lgbm_spec <- boost_tree(mode = "regression", min_n = tune()) %>%
    set_engine("lightgbm", objective = "rmse", verbose = -1,
               num_leaves = tune(), num_threads = 4L)

  lgbm_grid <- grid_max_entropy(
    parameters(min_n(c(2L, 60L), trans = NULL),
               num_leaves(c(2L, 256L))),
    size = 100
  )

  lgbm_wf <- workflow() %>%
    add_model(lgbm_spec) %>%
    add_formula(as.formula(paste(response_col, "~ .")))

  set.seed(913)
  lgbm_tuned <- tune_grid(lgbm_wf, resamples = tune_resamples,
                          grid = lgbm_grid, metrics = metric_set(rmse),
                          control = control_grid(verbose = FALSE))

  tune_scores <- lgbm_tuned %>%
    show_best(metric = "rmse", n = 100) %>%
    mutate(mean_rmse_rank = dense_rank(mean),
           se_rank        = dense_rank(std_err),
           combined_rank  = mean_rmse_rank + se_rank) %>%
    arrange(combined_rank)

  best_params <- tune_scores %>% slice(1)

  list(
    best_predictors = best_predictors,
    best_params     = best_params,
    varsel_stats    = varsel_stats,
    tune_scores     = tune_scores
  )
}
```
