---
title: "01_data_discovery — Daily p95 Turbidity Forecast (Modified Kemper)"
author: "ROSSyndicate"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(here)
library(tidyverse)
library(snotelr)
library(cdssr)
library(arrow)
```

# Purpose

Inventory all data sources needed for the daily p95 turbidity forecast workflow.
Confirms site availability, date coverage, and flags any gaps before data are
downloaded in `02_observational_data_prep.Rmd`.

Analogous to `kemper_turb_method/01_data_discovery.Rmd` but adapted for:
- Cache la Poudre watershed (CDSSR instead of USGS NWIS)
- Daily p95 turbidity target (pre-computed quantile CSV)
- SNOTEL snowpack network (new relative to Kemper)
- HEFS ensemble forecasts (NOAA NWPS API, analogous to Kemper's NWM)

---

# 1. Turbidity Target — FC Intake & Munroe Canal

```{r turb-inventory}
turb_raw <- read_csv(
  here("data/collated/FC_intake/fc_intake_munroe_turbidity_quantile_record_2010-2025.csv"),
  show_col_types = FALSE
)

# Date coverage and completeness by location
turb_raw %>%
  group_by(location) %>%
  summarise(
    first_date   = min(date),
    last_date    = max(date),
    n_total      = n(),
    n_complete   = sum(!is.na(p95)),
    pct_complete = round(100 * n_complete / n_total, 1)
  )
```

```{r turb-annual-gaps}
# Annual completeness for the intake (modeling target)
turb_raw %>%
  filter(location == "intake_turb_NTU") %>%
  mutate(year = lubridate::year(date)) %>%
  group_by(year) %>%
  summarise(
    n_days     = n(),
    n_complete = sum(!is.na(p95)),
    pct_complete = round(100 * n_complete / n_days, 1)
  ) %>%
  print(n = Inf)
```

**Note:** The `intake_turb_NTU` location is the modeling target.
`munroe_turb_NTU` has sparse coverage especially before 2018 and is not used
as a modeling target but may be useful as an upstream network predictor when
available.

---

# 2. Streamflow — Canyon Mouth (CLAFTCCO, CDSSR)

```{r flow-check}
# Pull a small test window to confirm API access and units
flow_test <- get_sw_ts(
  abbrev     = "CLAFTCCO",
  start_date = "2023-01-01",
  end_date   = "2023-12-31",
  timescale  = "day"
)

cat("Rows returned:", nrow(flow_test), "\n")
cat("Columns:", paste(names(flow_test), collapse = ", "), "\n")
cat("Flow units (parameter):", unique(flow_test$meas_unit), "\n")
cat("Date range:", as.character(range(flow_test$datetime)), "\n")
```

```{r flow-annual-summary}
# Full-period summary (will take a moment to pull)
flow_full <- get_sw_ts(
  abbrev     = "CLAFTCCO",
  start_date = "2010-01-01",
  end_date   = "2025-12-31",
  timescale  = "day"
)

flow_full %>%
  mutate(year = lubridate::year(datetime)) %>%
  group_by(year) %>%
  summarise(
    n_days     = n(),
    n_complete = sum(!is.na(value)),
    pct_complete = round(100 * n_complete / n_days, 1),
    mean_cfs   = round(mean(value, na.rm = TRUE), 1),
    max_cfs    = round(max(value,  na.rm = TRUE), 1)
  ) %>%
  print(n = Inf)
```

---

# 3. Precipitation — Redfeather Station (CDSSR climate)

```{r precip-check}
precip_test <- get_climate_ts(
  site_id    = "US1COLR0815",
  param      = "Precip",
  start_date = "2023-01-01",
  end_date   = "2023-12-31",
  timescale  = "day"
)

cat("Rows returned:", nrow(precip_test), "\n")
cat("Date range:", as.character(range(precip_test$datetime)), "\n")
cat("Units:", unique(precip_test$meas_unit), "\n")

# Annual completeness
get_climate_ts(
  site_id    = "US1COLR0815",
  param      = "Precip",
  start_date = "2010-01-01",
  end_date   = "2025-12-31",
  timescale  = "day"
) %>%
  mutate(year = lubridate::year(datetime)) %>%
  group_by(year) %>%
  summarise(
    n_days     = n(),
    n_complete = sum(!is.na(value)),
    pct_complete = round(100 * n_complete / n_days, 1)
  ) %>%
  print(n = Inf)
```

---

# 4. SNOTEL Network — Cache la Poudre HUC (10190007)

```{r snotel-sites}
# All SNOTEL sites that mention the CLP HUC in their description
snotel_sites <- snotel_info() %>%
  filter(grepl("10190007", description))

cat("SNOTEL sites in HUC 10190007:\n")
snotel_sites %>%
  select(site_id, site_name, state, start, end, description) %>%
  print()
```

```{r snotel-coverage}
# Download all sites to inspect coverage; filter to 2010–2025
snotel_all <- snotel_download(
  site_id  = snotel_sites$site_id,
  internal = TRUE,
  metric   = TRUE
) %>%
  mutate(date = lubridate::ymd(date),
         year = lubridate::year(date))

# SWE completeness per site per year
snotel_all %>%
  filter(lubridate::year(date) >= 2010) %>%
  group_by(site_id, year) %>%
  summarise(
    n_days     = n(),
    n_swe      = sum(!is.na(snow_water_equivalent)),
    pct_swe    = round(100 * n_swe / n_days, 1),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = pct_swe,
              names_prefix = "WY") %>%
  print()
```

```{r snotel-pca-preview}
# Quick PCA preview to confirm all sites load positively on PC1
snotel_matrix <- snotel_all %>%
  filter(lubridate::year(date) >= 2010) %>%
  select(date, site_id, snow_water_equivalent) %>%
  pivot_wider(names_from  = site_id,
              names_prefix = "swe_",
              values_from  = snow_water_equivalent) %>%
  filter(complete.cases(.))

pca_preview <- snotel_matrix %>%
  select(starts_with("swe_")) %>%
  prcomp(center = TRUE, scale. = TRUE)

cat("PC1 variance explained:",
    round(summary(pca_preview)$importance[2, 1] * 100, 1), "%\n")
cat("PC1 site loadings (all should be same sign):\n")
print(round(pca_preview$rotation[, 1], 3))
```

---

# 5. PRISM Air Temperature

```{r prism-check}
prism <- read_parquet(
  here("data/collated/external_datasets/daily_precip_temp_prism_2018-06-27_to_2025-08-25.parquet")
)

cat("PRISM columns:", paste(names(prism), collapse = ", "), "\n")
cat("Date range:", as.character(range(prism$date, na.rm = TRUE)), "\n")
cat("Rows:", nrow(prism), "\n")
glimpse(prism)
```

**Note:** PRISM temperature is available from 2018-06-27 onward only.
Temperature-based features (`tmax_5d_mean`, `swe_x_tmax`) will be NA for
2010–2018; LightGBM handles NA natively, but GAMs require complete rows.
Consider excluding those features from GAM formulas or using 2018+ data
for the temperature interaction term.

---

# 6. Larimer County Network Stations

```{r larimer-check}
larimer <- read_parquet(
  here("data/collated/external_datasets/larimer_county_station_flow_precip_20210101_20251201.parquet")
)

cat("Columns:", paste(names(larimer), collapse = ", "), "\n")
cat("Date range:", as.character(range(larimer$date, na.rm = TRUE)), "\n")

# Station inventory
larimer %>%
  count(station_name) %>%
  print(n = Inf)
```

**Note:** Larimer County network data begins 2021-01-01. These stations can
serve as an upstream network analog to Kemper's tributary stations, but
coverage is too short for model training (which uses 2010–2023). They may
be used as supplemental predictors in an extended model or for operational
HEFS feature construction (scripts 06–07).

---

# 7. HEFS Forecast Availability — Canyon Mouth Gauge

```{r hefs-gauge-check}
source(here("src/hefs/get_hefs_gauges.R"))

# Find the NWPS gauge ID corresponding to Canyon Mouth / Poudre at Canyon Mouth
# Typically searched by state or HUC; adjust state as needed
hefs_gauges <- get_hefs_gauges()

# Filter to Colorado Poudre-area gauges
hefs_gauges %>%
  filter(grepl("POUDRE|CANYON|COLOC", lid, ignore.case = TRUE) |
           grepl("Colorado", state, ignore.case = TRUE)) %>%
  select(lid, name, state, rfc) %>%
  print()
```

```{r hefs-test-pull}
source(here("src/hefs/get_hefs_ensembles.R"))

# Test pull using the confirmed gauge ID (update CANYON_MOUTH_ID after inspection above)
# The NWPS gauge for Canyon Mouth is typically "COLOC070" or similar — confirm above.
CANYON_MOUTH_HEFS_ID <- "COLOC070"  # update after running hefs-gauge-check

test_forecast <- get_hefs_ensemble(
  ID   = CANYON_MOUTH_HEFS_ID,
  date = Sys.Date() - 1  # yesterday's forecast
)

if (!is.null(test_forecast)) {
  cat("HEFS test pull successful.\n")
  cat("Forecast rows:", nrow(test_forecast), "\n")
  cat("Date range:", as.character(range(test_forecast$date_time)), "\n")
  cat("Ensemble members:", length(unique(test_forecast$ensemble_member)), "\n")
  glimpse(test_forecast)
} else {
  cat("HEFS pull returned NULL — check gauge ID and API availability.\n")
}
```

---

# Summary

```{r summary-table}
tibble::tribble(
  ~Data_Source,        ~Coverage,            ~Resolution, ~Status,
  "Turbidity (intake)", "2010–2025",          "daily p05–p95", "ready",
  "Canyon Mouth Q",    "2010–2025",          "daily CFS",     "ready",
  "Redfeather Precip", "2010–2025",          "daily inches",  "ready",
  "SNOTEL SWE",        "2010–2025 (varies)", "daily mm",      "confirm coverage",
  "PRISM Tmax/Tmin",   "2018–2025",          "daily °C",      "ready (2018+)",
  "Larimer County Q/P","2021–2025",          "daily",         "supplemental only",
  "HEFS Forecasts",    "2023–2025+",         "ensemble daily","confirm gauge ID"
) %>% knitr::kable()
```
