---
title: "02_observational_data_prep — Daily p95 Turbidity Forecast (Modified Kemper)"
author: "ROSSyndicate"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(here)
library(tidyverse)
library(lubridate)
library(snotelr)
library(cdssr)
library(arrow)
library(RcppRoll)
```

# Purpose

Download and clean all observational data used as predictors. Outputs a single
clean parquet file (`daily_obs_drivers_clean.parquet`) that is the starting
point for `03_feature_engineering.Rmd`.

Adaptations from the original Kemper `02_observational_data_download_and_clean.Rmd`:
- Target is daily p95 turbidity (NTU) from a pre-computed CSV — no sub-hourly
  aggregation or probe-conversion is required.
- Discharge is daily CFS from CDSSR (single station; no tributary network).
- Snow/melt and temperature predictors are added for the CLP environment.
- No tunnel-flow correction is applied.

---

# 1. Turbidity Target — FC Intake

```{r load-turbidity}
turb <- read_csv(
  here("data/collated/FC_intake/fc_intake_munroe_turbidity_quantile_record_2010-2025.csv"),
  show_col_types = FALSE
) %>%
  filter(location == "intake_turb_NTU") %>%
  mutate(date = as.Date(date)) %>%
  select(date, doy, year, p05, p25, p50, p75, p95, n_obs = n)

cat("Turbidity rows:", nrow(turb), "\n")
cat("Date range:", as.character(range(turb$date)), "\n")
cat("Complete p95 rows:", sum(!is.na(turb$p95)), "\n")
```

---

# 2. Streamflow — Canyon Mouth (CLAFTCCO, CDSSR)

```{r load-flow}
flow_raw <- get_sw_ts(
  abbrev     = "CLAFTCCO",
  start_date = "2010-01-01",
  end_date   = "2025-12-31",
  timescale  = "day"
) %>%
  mutate(date = as.Date(datetime)) %>%
  select(date, canyon_mouth_cfs = value)

cat("Flow rows:", nrow(flow_raw), "\n")
cat("NA flow values:", sum(is.na(flow_raw$canyon_mouth_cfs)), "\n")
```

```{r flow-gap-fill}
# Fill gaps up to 3 consecutive days via linear interpolation
# (longer gaps left as NA for the model to handle)
flow_clean <- flow_raw %>%
  arrange(date) %>%
  mutate(
    canyon_mouth_cfs = zoo::na.approx(canyon_mouth_cfs,
                                      maxgap = 3,
                                      na.rm  = FALSE)
  )

cat("NA flow values after gap-fill (<=3d):", sum(is.na(flow_clean$canyon_mouth_cfs)), "\n")
```

---

# 3. Precipitation — Redfeather Station (US1COLR0815, CDSSR)

```{r load-precip}
precip_raw <- get_climate_ts(
  site_id    = "US1COLR0815",
  param      = "Precip",
  start_date = "2010-01-01",
  end_date   = "2025-12-31",
  timescale  = "day"
) %>%
  mutate(date = as.Date(datetime)) %>%
  select(date, redfeather_precip_in = value)

cat("Precip rows:", nrow(precip_raw), "\n")
cat("NA precip values:", sum(is.na(precip_raw$redfeather_precip_in)), "\n")
```

---

# 4. SNOTEL SWE — Cache la Poudre HUC (10190007)

```{r snotel-download}
snotel_sites <- snotel_info() %>%
  filter(grepl("10190007", description))

cat("Downloading SWE for", nrow(snotel_sites), "SNOTEL sites...\n")

snotel_raw <- snotel_download(
  site_id  = snotel_sites$site_id,
  internal = TRUE,
  metric   = TRUE
) %>%
  mutate(date = ymd(date))
```

```{r snotel-pca}
# PCA on SWE across sites — PC1 captures dominant snowpack signal
# Restrict PCA to dates with complete observations across all sites
snotel_matrix <- snotel_raw %>%
  filter(year(date) >= 2010) %>%
  select(date, site_id, snow_water_equivalent) %>%
  pivot_wider(names_from  = site_id,
              names_prefix = "swe_",
              values_from  = snow_water_equivalent) %>%
  filter(complete.cases(.))

pca_snotel <- snotel_matrix %>%
  select(starts_with("swe_")) %>%
  prcomp(center = TRUE, scale. = TRUE)

cat("PC1 variance explained:",
    round(summary(pca_snotel)$importance[2, 1] * 100, 1), "%\n")

# Confirm sign convention: all loadings should be positive
# If any are negative, flip the sign of swe_pc1 so that:
#   high swe_pc1 = high snowpack, melt_rate_7d > 0 = active melt
cat("PC1 site loadings:\n")
print(round(pca_snotel$rotation[, 1], 3))

snotel <- tibble(
  date    = snotel_matrix$date,
  swe_pc1 = pca_snotel$x[, 1]
) %>%
  arrange(date)

# If PC1 loads negatively (all loadings < 0), flip sign
if (all(pca_snotel$rotation[, 1] < 0)) {
  snotel <- snotel %>% mutate(swe_pc1 = -swe_pc1)
  message("PC1 sign flipped: loadings were all negative.")
}
```

---

# 5. PRISM Air Temperature (2018–2025)

```{r load-temp}
prism <- read_parquet(
  here("data/collated/external_datasets/daily_precip_temp_prism_2018-06-27_to_2025-08-25.parquet")
)

# Inspect columns — may be gridded (long format) or already aggregated
glimpse(prism)

# Aggregate to watershed-mean Tmax and Tmin if gridded
if ("tmax" %in% names(prism) || any(grepl("tmax", names(prism), ignore.case = TRUE))) {
  tmax_col <- grep("tmax", names(prism), ignore.case = TRUE, value = TRUE)[1]
  tmin_col <- grep("tmin", names(prism), ignore.case = TRUE, value = TRUE)[1]

  prism_daily <- prism %>%
    mutate(date = as.Date(date)) %>%
    group_by(date) %>%
    summarise(
      tmax_c = mean(.data[[tmax_col]], na.rm = TRUE),
      tmin_c = mean(.data[[tmin_col]], na.rm = TRUE),
      .groups = "drop"
    )
} else {
  # Already daily; rename as needed
  prism_daily <- prism %>%
    mutate(date = as.Date(date)) %>%
    rename_with(tolower) %>%
    select(date, contains("tmax"), contains("tmin"))
}

cat("PRISM temperature rows:", nrow(prism_daily), "\n")
cat("Date range:", as.character(range(prism_daily$date)), "\n")
```

---

# 6. Assemble Clean Dataset

```{r assemble}
# Build a complete daily spine 2010-01-01 to 2025-12-31
date_spine <- tibble(date = seq.Date(as.Date("2010-01-01"),
                                     as.Date("2025-12-31"),
                                     by = "day"))

daily_obs_drivers_clean <- date_spine %>%
  left_join(turb    %>% select(date, doy, year, p95), by = "date") %>%
  left_join(flow_clean,                                by = "date") %>%
  left_join(precip_raw,                                by = "date") %>%
  left_join(snotel,                                    by = "date") %>%
  left_join(prism_daily,                               by = "date") %>%
  mutate(
    doy  = if_else(is.na(doy),  yday(date),  doy),
    year = if_else(is.na(year), year(date),  year)
  ) %>%
  arrange(date)

cat("Assembled dataset rows:", nrow(daily_obs_drivers_clean), "\n")

# NA counts per column
daily_obs_drivers_clean %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "n_na") %>%
  filter(n_na > 0) %>%
  mutate(pct_na = round(100 * n_na / nrow(daily_obs_drivers_clean), 1)) %>%
  arrange(desc(n_na)) %>%
  print()
```

---

# 7. Save

```{r save}
out_dir <- here("data/derived/models/turbidity_forecast")
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

write_parquet(
  daily_obs_drivers_clean,
  file.path(out_dir, "daily_obs_drivers_clean.parquet")
)

cat("Saved:", file.path(out_dir, "daily_obs_drivers_clean.parquet"), "\n")
```
