---
title: "03_feature_engineering — Daily p95 Turbidity Forecast (Modified Kemper)"
author: "ROSSyndicate"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(here)
library(tidyverse)
library(lubridate)
library(slider)
library(arrow)

# Source helper functions
source(knitr::purl(here("modeling/turbidity/modified_kemper_application_daily/00_functions.Rmd"),
                   quiet = TRUE))
```

# Purpose

Build the complete modelling dataset from `daily_obs_drivers_clean.parquet`.
Computes all predictor features (flow antecedents, snow/melt, precipitation,
temperature interaction) and the regime classification, then creates 1-, 2-,
and 3-day lead targets.

Outputs `daily_model_data.parquet`.

---

# Load Clean Drivers

```{r load-data}
obs <- read_parquet(
  here("data/derived/models/turbidity_forecast/daily_obs_drivers_clean.parquet")
)

cat("Rows:", nrow(obs), "\n")
cat("Date range:", as.character(range(obs$date)), "\n")
glimpse(obs)
```

---

# Flow Features

```{r flow-features}
# antecedent_Q_maker() defined in 00_functions.Rmd
model_data <- antecedent_Q_maker(obs)

cat("Flow feature columns added:\n")
cat(setdiff(names(model_data), names(obs)), "\n")
```

---

# Snow / Melt Features

```{r snow-features}
# melt_rate_calculator() defined in 00_functions.Rmd
model_data <- melt_rate_calculator(model_data)

cat("NA melt_rate_7d:", sum(is.na(model_data$melt_rate_7d)),
    "— expected for first 7 rows and any SNOTEL gaps\n")
```

---

# Precipitation Features

```{r precip-features}
model_data <- model_data %>%
  arrange(date) %>%
  mutate(
    # Treat NA precipitation as 0 (trace / missing = no event assumed)
    precip        = replace_na(redfeather_precip_in, 0),

    precip_3day   = slide_dbl(precip, sum, .before = 2, .complete = TRUE),
    precip_7day   = slide_dbl(precip, sum, .before = 6, .complete = TRUE),

    precip_event  = as.integer(precip > PRECIP_THRESH)
  )
```

---

# Temperature Interaction Feature (2018+ only)

```{r temp-features}
# 5-day rolling mean Tmax and SWE × Tmax interaction term
# These columns will be NA before 2018-06-27; LightGBM handles NA natively.
# Exclude from GAM formulas or restrict to 2018+ when fitting GAMs with temp terms.

model_data <- model_data %>%
  arrange(date) %>%
  mutate(
    tmax_5d_mean = slide_dbl(tmax_c, mean, .before = 4, .complete = TRUE),
    swe_x_tmax   = swe_pc1 * tmax_5d_mean
  )
```

---

# Regime Classification

```{r regime-classification}
# regime_classifier() defined in 00_functions.Rmd
model_data <- regime_classifier(model_data)

# Check distribution
model_data %>%
  filter(year %in% 2010:2023) %>%
  count(regime) %>%
  mutate(pct = round(100 * n / sum(n), 1))
```

---

# Lead Targets (t+1, t+2, t+3)

```{r lead-targets}
model_data <- model_data %>%
  arrange(date) %>%
  mutate(
    target_p95_t1 = lead(p95, 1),
    target_p95_t2 = lead(p95, 2),
    target_p95_t3 = lead(p95, 3)
  )
```

---

# Verify Completeness

```{r completeness}
# Columns required to fit any model (GAM-safe subset — excludes temp interaction)
MODEL_COLS <- c("p95", "log_q", "deltaQ_24h",
                "precip_3day", "precip_event",
                "melt_rate_7d", "doy", "regime")

TARGET_COLS <- c("target_p95_t1", "target_p95_t2", "target_p95_t3")

cat("Complete rows (MODEL_COLS + t+1 target):",
    sum(complete.cases(model_data[, c(MODEL_COLS, "target_p95_t1")])),
    "of", nrow(model_data), "\n")

# Regime distribution for training years only
model_data %>%
  filter(year %in% 2010:2023,
         complete.cases(.[, c(MODEL_COLS, "target_p95_t1")])) %>%
  count(regime) %>%
  mutate(pct = round(100 * n / sum(n), 1))
```

---

# Regime Visualisation

```{r regime-ts, fig.dim = c(12, 4)}
model_data %>%
  filter(year %in% 2010:2023,
         complete.cases(.[, c(MODEL_COLS, "target_p95_t1")])) %>%
  ggplot(aes(x = date, y = p95, color = regime)) +
  geom_point(size = 0.4, alpha = 0.6) +
  scale_color_manual(
    values = c(snowmelt = "#2166ac", convective = "#d73027", baseflow = "#4dac26")
  ) +
  scale_y_continuous(limits = c(0, NA)) +
  theme_bw() +
  labs(title = "Observed daily p95 turbidity coloured by regime (2010–2023)",
       x = NULL, y = "p95 turbidity (NTU)")
```

```{r regime-boxplot, fig.dim = c(6, 5)}
model_data %>%
  filter(year %in% 2010:2023,
         complete.cases(.[, c(MODEL_COLS, "target_p95_t1")])) %>%
  ggplot(aes(x = regime, y = p95, fill = regime)) +
  geom_boxplot(outlier.size = 0.5) +
  scale_fill_manual(
    values = c(snowmelt = "#2166ac", convective = "#d73027", baseflow = "#4dac26")
  ) +
  scale_y_log10() +
  theme_bw() +
  theme(legend.position = "none") +
  labs(title = "Daily p95 turbidity by regime (log scale)", x = NULL, y = "NTU (log)")
```

---

# Antecedent Flow — Seasonal Profile

```{r antecedent-profile, fig.dim = c(10, 4)}
# Visualise seasonal pattern of key antecedent features
model_data %>%
  filter(year %in% 2010:2023) %>%
  group_by(doy) %>%
  summarise(
    median_q       = median(canyon_mouth_cfs, na.rm = TRUE),
    median_melt7d  = median(melt_rate_7d,     na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(median_q, median_melt7d),
               names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = doy, y = value)) +
  geom_line() +
  facet_wrap(~ variable, scales = "free_y") +
  theme_bw() +
  labs(title = "Median seasonal profiles — discharge and melt rate (2010–2023)",
       x = "Day of year", y = NULL)
```

---

# Save

```{r save}
out_dir <- here("data/derived/models/turbidity_forecast")
write_parquet(model_data,
              file.path(out_dir, "daily_model_data.parquet"))

cat("Saved:", file.path(out_dir, "daily_model_data.parquet"), "\n")
cat("Columns:", paste(names(model_data), collapse = ", "), "\n")
```
