---
title: "04_event_delineation — Daily p95 Turbidity Forecast (Modified Kemper)"
author: "ROSSyndicate"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(here)
library(tidyverse)
library(lubridate)
library(arrow)
```

# Purpose

Identify and characterise high-turbidity events in the historical record.
Produces an **event catalog** (`turbidity_event_catalog.parquet`) that is used in
`09_evaluate_performance.Rmd` for event-based forecast evaluation.

Analogous to `kemper_turb_method/04_event_delineation.Rmd` but adapted for:
- Daily resolution (no hydroEvents package needed — events defined by threshold runs)
- Two distinct event types: **snowmelt** (DOY 90–180) and **convective** (DOY 150–300)
- NTU p95 target rather than instantaneous FNU

---

# Constants

```{r constants}
# Turbidity threshold above which a day is considered an "event day"
# Adjust after visual inspection of the distribution
TURB_EVENT_THRESH <- 10   # NTU p95: elevated, not extreme
TURB_HIGH_THRESH  <- 75   # NTU p95: high-turbidity event (treatment concern)

# Minimum number of consecutive event days to count as an event
MIN_EVENT_DAYS <- 1

# DOY windows for regime-based event typing
SNOWMELT_DOY_RANGE  <- 90:180
CONVECTIVE_DOY_RANGE <- 150:300
```

---

# Load Data

```{r load-data}
model_data <- read_parquet(
  here("data/derived/models/turbidity_forecast/daily_model_data.parquet")
)

# Restrict to training period for event catalog
train_data <- model_data %>%
  filter(year %in% 2010:2023, !is.na(p95))

cat("Training rows with observed p95:", nrow(train_data), "\n")
```

---

# Turbidity Distribution

```{r turb-distribution, fig.dim = c(8, 4)}
train_data %>%
  ggplot(aes(x = p95)) +
  geom_histogram(bins = 60, fill = "steelblue", color = "white") +
  scale_x_log10() +
  geom_vline(xintercept = TURB_EVENT_THRESH, color = "orange",
             linetype = "dashed", linewidth = 0.8) +
  geom_vline(xintercept = TURB_HIGH_THRESH,  color = "red",
             linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = TURB_EVENT_THRESH * 1.2, y = Inf,
           label = paste0("Event (", TURB_EVENT_THRESH, " NTU)"),
           hjust = 0, vjust = 1.5, color = "orange", size = 3) +
  annotate("text", x = TURB_HIGH_THRESH * 1.2, y = Inf,
           label = paste0("High (", TURB_HIGH_THRESH, " NTU)"),
           hjust = 0, vjust = 1.5, color = "red", size = 3) +
  theme_bw() +
  labs(title = "Distribution of daily p95 turbidity (2010–2023, log scale)",
       x = "p95 NTU", y = "Count")
```

```{r percentile-summary}
# How many days exceed each threshold?
train_data %>%
  summarise(
    n_above_event = sum(p95 > TURB_EVENT_THRESH, na.rm = TRUE),
    n_above_high  = sum(p95 > TURB_HIGH_THRESH,  na.rm = TRUE),
    pct_above_event = round(100 * n_above_event / n(), 1),
    pct_above_high  = round(100 * n_above_high  / n(), 1)
  )
```

---

# Event Detection

Events are defined as contiguous runs of days where `p95 > TURB_EVENT_THRESH`.
Each run is assigned a unique event ID, and its regime type is determined by
the majority regime during the event.

```{r event-detection}
events_raw <- train_data %>%
  arrange(date) %>%
  mutate(
    is_event = p95 > TURB_EVENT_THRESH,
    # Assign run-length IDs; an event starts when is_event switches from FALSE to TRUE
    run_id   = cumsum(!is_event | (is_event & lag(!is_event, default = TRUE)))
  ) %>%
  filter(is_event) %>%
  group_by(run_id) %>%
  summarise(
    event_start    = min(date),
    event_end      = max(date),
    event_duration = as.integer(difftime(max(date), min(date), units = "days")) + 1L,
    peak_p95       = max(p95, na.rm = TRUE),
    mean_p95       = mean(p95, na.rm = TRUE),
    peak_q_cfs     = max(canyon_mouth_cfs, na.rm = TRUE),
    mean_q_cfs     = mean(canyon_mouth_cfs, na.rm = TRUE),
    peak_doy       = doy[which.max(p95)],
    majority_regime = names(sort(table(regime), decreasing = TRUE))[1],
    is_high_turb   = any(p95 > TURB_HIGH_THRESH),
    .groups = "drop"
  ) %>%
  filter(event_duration >= MIN_EVENT_DAYS)

cat("Events detected:", nrow(events_raw), "\n")
cat("High-turbidity events (>", TURB_HIGH_THRESH, "NTU):",
    sum(events_raw$is_high_turb), "\n")
```

```{r event-typing}
# Classify event type based on peak DOY and majority regime
turbidity_events <- events_raw %>%
  mutate(
    event_type = case_when(
      majority_regime == "snowmelt"   ~ "snowmelt",
      majority_regime == "convective" ~ "convective",
      .default = "baseflow/other"
    ),
    event_type = factor(event_type,
                        levels = c("snowmelt", "convective", "baseflow/other"))
  ) %>%
  mutate(event_id = row_number(), .before = event_start)

turbidity_events %>%
  count(event_type) %>%
  mutate(pct = round(100 * n / sum(n), 1))
```

---

# Event Summary by Year

```{r events-per-year}
turbidity_events %>%
  mutate(year = year(event_start)) %>%
  count(year, event_type) %>%
  pivot_wider(names_from = event_type, values_from = n, values_fill = 0) %>%
  print(n = Inf)
```

---

# Event Timeline

```{r event-timeline, fig.dim = c(12, 5)}
train_data %>%
  left_join(
    turbidity_events %>%
      select(event_start, event_end, event_type) %>%
      mutate(date = map2(event_start, event_end, seq.Date, by = "day")) %>%
      unnest(date) %>%
      select(date, event_type),
    by = "date"
  ) %>%
  ggplot(aes(x = date, y = p95)) +
  geom_line(color = "grey60", linewidth = 0.3) +
  geom_point(aes(color = event_type), size = 0.8, na.rm = TRUE) +
  scale_color_manual(
    values = c(snowmelt = "#2166ac", convective = "#d73027", "baseflow/other" = "#b2df8a"),
    na.value = "grey80"
  ) +
  geom_hline(yintercept = TURB_EVENT_THRESH, linetype = "dashed",
             color = "orange", linewidth = 0.5) +
  geom_hline(yintercept = TURB_HIGH_THRESH,  linetype = "dashed",
             color = "red",    linewidth = 0.5) +
  scale_y_log10() +
  theme_bw() +
  labs(title = "Daily p95 turbidity — event periods coloured by type (2010–2023)",
       x = NULL, y = "p95 NTU (log scale)", color = "Event type")
```

---

# Seasonal Event Distribution

```{r seasonal-distribution, fig.dim = c(8, 4)}
turbidity_events %>%
  ggplot(aes(x = peak_doy, fill = event_type)) +
  geom_histogram(bins = 36, position = "stack") +
  scale_fill_manual(
    values = c(snowmelt = "#2166ac", convective = "#d73027", "baseflow/other" = "#b2df8a")
  ) +
  theme_bw() +
  labs(title = "Seasonal distribution of turbidity events (peak DOY)",
       x = "Day of year (event peak)", y = "Count", fill = "Event type")
```

---

# Save Event Catalog

```{r save}
out_dir <- here("data/derived/models/turbidity_forecast")
write_parquet(turbidity_events,
              file.path(out_dir, "turbidity_event_catalog.parquet"))

cat("Saved:", file.path(out_dir, "turbidity_event_catalog.parquet"), "\n")
cat("Events:", nrow(turbidity_events), "\n")
```
