---
title: "05_model_development — Daily p95 Turbidity Forecast (Modified Kemper)"
author: "ROSSyndicate"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(here)
library(tidyverse)
library(lubridate)
library(arrow)
library(mgcv)
library(quantreg)
library(hydroGOF)
library(lightgbm)
library(tidymodels)
library(bonsai)

# Source helper functions
source(knitr::purl(here("modeling/turbidity/modified_kemper_application_daily/00_functions.Rmd"),
                   quiet = TRUE))
```

# Purpose

Develop and cross-validate two model families for daily p95 turbidity forecasting
at the FC Intake. Following the Kemper et al. dual-model approach:

1. **Hierarchical GAM** — three regime-specific GAMs (snowmelt, convective,
   baseflow) plus a global GAM benchmark. Adapted from `intake_turbidity_forecast_poc.Rmd`.

2. **LightGBM** — gradient-boosted model with full antecedent feature set.
   Adapted from `kemper_application/02_model_development.Rmd`.

Models are fit for each of three lead times (t+1, t+2, t+3).

**Train/CV/test split:**
- Training + CV: 2010–2023 (5-fold year-grouped CV)
- Test (held out): 2024–2025

---

# Constants

```{r constants}
TRAIN_CV_YEARS <- 2010:2023
TEST_YEARS     <- 2024:2025
HIGH_TURB_THRESH <- 75  # NTU — threshold for high-turbidity event evaluation

# 5 grouped-year CV folds
cv_folds <- list(
  fold_1 = 2010:2012,
  fold_2 = 2013:2015,
  fold_3 = 2016:2018,
  fold_4 = 2019:2020,
  fold_5 = 2021:2023
)

# Columns required for any model (GAM-safe: no temperature interaction)
MODEL_COLS <- c("p95", "log_q", "deltaQ_24h",
                "precip_3day", "precip_event",
                "melt_rate_7d", "doy", "regime")

# All candidate LightGBM predictors (superset; includes temp features for 2018+)
LGBM_CANDIDATE_PREDS <- c(
  "log_q", "log_q_sq", "log_limb", "dq_q",
  "deltaQ_24h", "deltaQ_48h",
  "mean_q_7d", "mean_q_30d", "mean_q_90d",
  "max_q_30d", "max_q_90d", "max_q_365d",
  "time_since_30d_peak", "time_since_90d_peak", "time_since_annual_peak",
  "swe_pc1", "melt_rate_7d", "melt_rate_14d",
  "precip", "precip_3day", "precip_7day", "precip_event",
  "tmax_5d_mean", "swe_x_tmax",
  "doy", "regime"
)

# GAM formulas — Gamma(link="log"); target column renamed to `target_p95` in helper
GLOBAL_FORMULA <- target_p95 ~
  s(doy, bs = "cc", k = 10) +
  s(log_q, k = 10) +
  s(deltaQ_24h, k = 8) +
  precip_event +
  s(melt_rate_7d, bs = "ts", k = 8)

REGIME_FORMULAS <- list(
  snowmelt   = target_p95 ~
    s(log_q, k = 10) +
    s(melt_rate_7d, bs = "ts", k = 8),
  convective = target_p95 ~
    s(log_q, k = 8) +
    s(deltaQ_24h, k = 8) +
    precip_event,
  baseflow   = target_p95 ~
    s(doy, bs = "cc", k = 10) +
    s(log_q, k = 8)
)
```

---

# Load Data

```{r load-data}
model_data <- read_parquet(
  here("data/derived/models/turbidity_forecast/daily_model_data.parquet")
) %>%
  mutate(regime = factor(regime, levels = c("snowmelt", "convective", "baseflow")))

cat("Total rows:", nrow(model_data), "\n")

train_cv_data <- model_data %>% filter(year %in% TRAIN_CV_YEARS)
test_data      <- model_data %>% filter(year %in% TEST_YEARS)

cat("Train+CV rows:", nrow(train_cv_data), "| Test rows:", nrow(test_data), "\n")
```

---

# Model A: Hierarchical GAM

## Cross-Validation

```{r gam-cv}
# run_gam_cv_fold() defined in 00_functions.Rmd
# Run CV for each lead time (1, 2, 3) and bind results

gam_cv_preds <- map_dfr(1:3, function(lead) {
  map_dfr(names(cv_folds), function(k) {
    run_gam_cv_fold(
      k              = k,
      model_data     = train_cv_data,
      cv_folds       = cv_folds,
      TRAIN_CV_YEARS = TRAIN_CV_YEARS,
      MODEL_COLS     = MODEL_COLS,
      GLOBAL_FORMULA = GLOBAL_FORMULA,
      REGIME_FORMULAS = REGIME_FORMULAS,
      lead           = lead
    )
  })
})

cat("CV prediction rows:", nrow(gam_cv_preds), "\n")
```

```{r gam-cv-metrics}
# Compute performance metrics for each model × lead time
gam_cv_metrics <- gam_cv_preds %>%
  drop_na(observed, global_pred, regime_pred, persistence) %>%
  group_by(lead) %>%
  summarise(
    persistence = error_estimator(persistence, observed, "persistence")$rmse,
    global_gam  = error_estimator(global_pred,  observed, "global")$rmse,
    regime_gam  = error_estimator(regime_pred,  observed, "regime")$rmse,
    .groups = "drop"
  )

cat("GAM cross-validation RMSE by lead time:\n")
print(gam_cv_metrics)
```

## Fit Final GAMs on Full Training Period (2010–2023)

```{r gam-final-fit}
# Final models retrained on ALL 2010–2023 data for test-period evaluation
final_gam_models <- map(1:3, function(lead) {
  target_col <- paste0("target_p95_t", lead)

  train_full <- train_cv_data %>%
    rename(target_p95 = !!target_col) %>%
    drop_na(all_of(c(MODEL_COLS, "target_p95")))

  list(
    global = gam(GLOBAL_FORMULA, data = train_full,
                 family = Gamma(link = "log"), method = "REML"),
    regime = map(REGIME_FORMULAS, function(f) {
      gam(f, data = train_full, family = Gamma(link = "log"), method = "REML")
    })
  )
})
names(final_gam_models) <- paste0("lead_t", 1:3)

cat("Final GAMs fitted for leads t+1, t+2, t+3.\n")
```

```{r gam-summary}
# Brief summary of each global GAM
map(names(final_gam_models), function(nm) {
  cat("\n---", nm, "Global GAM ---\n")
  print(summary(final_gam_models[[nm]]$global))
})
```

---

# Model B: LightGBM

LightGBM is fit separately for each lead time. Variable selection and
hyperparameter tuning use expanding-window CV (adapts `lgbm_tuner_daily()`
from `00_functions.Rmd`, which mirrors `kemper_application/02_model_development.Rmd`).

## Data Preparation — Log-Transform Target

```{r lgbm-data-prep}
# LightGBM is trained on log10(p95) to compress the right tail
# Regime is converted to integer for LightGBM (1 = snowmelt, 2 = convective, 3 = baseflow)
lgbm_base <- train_cv_data %>%
  filter(!is.na(p95), p95 > 0) %>%
  mutate(
    log_p95_t1 = log10(lead(p95, 1)),
    log_p95_t2 = log10(lead(p95, 2)),
    log_p95_t3 = log10(lead(p95, 3)),
    regime_int = as.integer(regime)
  ) %>%
  # Replace regime factor with integer for LightGBM
  mutate(regime = regime_int) %>%
  select(-regime_int)
```

## Variable Selection + Hyperparameter Tuning

```{r lgbm-varsel-tune, eval=FALSE}
# NOTE: eval=FALSE — this is computationally expensive.
# Run once and save results; reload from cache below.

lgbm_tuning_results <- map(1:3, function(lead) {
  response_col <- paste0("log_p95_t", lead)
  cat("\n=== LightGBM variable selection: lead t+", lead, "===\n")

  lgbm_tuner_daily(
    df           = lgbm_base,
    all_preds    = LGBM_CANDIDATE_PREDS,
    response_col = response_col
  )
})
names(lgbm_tuning_results) <- paste0("lead_t", 1:3)

# Save tuning results
saveRDS(lgbm_tuning_results,
        here("data/derived/models/turbidity_forecast/lgbm_tuning_results.rds"))
```

```{r lgbm-load-tuning}
# Load previously cached tuning results
lgbm_tuning_results <- readRDS(
  here("data/derived/models/turbidity_forecast/lgbm_tuning_results.rds")
)

# Report selected predictors and hyperparameters per lead time
map(names(lgbm_tuning_results), function(nm) {
  res <- lgbm_tuning_results[[nm]]
  cat("\n---", nm, "---\n")
  cat("Selected predictors:", paste(res$best_predictors, collapse = ", "), "\n")
  cat("Best min_n:", res$best_params$min_n,
      "| Best num_leaves:", res$best_params$num_leaves, "\n")
})
```

## Train Final LightGBM Models

```{r lgbm-final-train, eval=FALSE}
final_lgbm_models <- map(1:3, function(lead) {
  response_col <- paste0("log_p95_t", lead)
  res          <- lgbm_tuning_results[[paste0("lead_t", lead)]]

  train_d <- lgbm_base %>%
    filter(year %in% TRAIN_CV_YEARS) %>%
    select(all_of(c(response_col, res$best_predictors))) %>%
    filter(is.finite(.data[[response_col]]))

  train_mat   <- data.matrix(train_d %>% select(!all_of(response_col)))
  train_label <- train_d[[response_col]]
  lgbm_ds     <- lgb.Dataset(train_mat, label = train_label)

  final_params <- list(
    objective        = "regression",
    num_leaves       = res$best_params$num_leaves,
    min_data_in_leaf = res$best_params$min_n,
    learning_rate    = 0.1,
    num_threads      = 4L,
    verbose          = -1L
  )

  set.seed(913)
  lgbm_mod <- lgb.train(final_params, data = lgbm_ds, nrounds = 100L)

  # Smearing factor (Duan 1983)
  smear_d <- mean(10^(train_label - predict(lgbm_mod, data = train_mat)),
                  na.rm = TRUE)

  cat("Lead t+", lead, "smearing factor:", round(smear_d, 4), "\n")

  list(model   = lgbm_mod,
       smear_d = smear_d,
       preds   = res$best_predictors)
})
names(final_lgbm_models) <- paste0("lead_t", 1:3)

# Save models
map(names(final_lgbm_models), function(nm) {
  lgb.save(
    final_lgbm_models[[nm]]$model,
    filename = here("data/derived/models/turbidity_forecast",
                    paste0("lgbm_daily_", nm, ".txt"))
  )
})

# Save smearing factors and predictor lists
lgbm_model_meta <- map(final_lgbm_models,
                       ~ list(smear_d = .x$smear_d, preds = .x$preds))
saveRDS(lgbm_model_meta,
        here("data/derived/models/turbidity_forecast/lgbm_model_meta.rds"))
```

---

# Test Period Evaluation

```{r test-eval}
test_input <- test_data %>%
  mutate(
    log_p95_t1 = log10(lead(p95, 1)),
    log_p95_t2 = log10(lead(p95, 2)),
    log_p95_t3 = log10(lead(p95, 3)),
    regime_int = as.integer(regime)
  ) %>%
  mutate(regime = regime_int) %>%
  select(-regime_int)

lgbm_model_meta <- readRDS(
  here("data/derived/models/turbidity_forecast/lgbm_model_meta.rds")
)

all_test_results <- map_dfr(1:3, function(lead) {
  response_col <- paste0("log_p95_t", lead)
  meta         <- lgbm_model_meta[[paste0("lead_t", lead)]]

  lgbm_mod <- lgb.load(
    filename = here("data/derived/models/turbidity_forecast",
                    paste0("lgbm_daily_lead_t", lead, ".txt"))
  )

  test_d   <- test_input %>%
    select(date, year, doy, regime, p95, all_of(response_col),
           all_of(meta$preds)) %>%
    filter(is.finite(.data[[response_col]]))
  test_mat <- data.matrix(test_d %>% select(all_of(meta$preds)))

  lgbm_pred <- (10^predict(lgbm_mod, data = test_mat)) * meta$smear_d

  # GAM predictions
  gam_final <- final_gam_models[[paste0("lead_t", lead)]]
  gam_global <- as.numeric(predict(gam_final$global,
                                   newdata = test_d, type = "response"))
  gam_regime <- map2_dbl(
    as.character(test_d$regime), seq_len(nrow(test_d)),
    ~ as.numeric(predict(gam_final$regime[[.x]],
                         newdata = test_d[.y, ], type = "response"))
  )

  tibble(
    date        = test_d$date,
    lead        = lead,
    regime      = test_d$regime,
    observed    = 10^test_d[[response_col]],
    persistence = test_d$p95,
    gam_global  = gam_global,
    gam_regime  = gam_regime,
    lgbm        = lgbm_pred
  )
})
```

```{r test-metrics}
all_test_results %>%
  drop_na() %>%
  group_by(lead) %>%
  summarise(
    persistence = error_estimator(persistence, observed, "persistence")$rmse,
    gam_global  = error_estimator(gam_global,  observed, "gam_global")$rmse,
    gam_regime  = error_estimator(gam_regime,  observed, "gam_regime")$rmse,
    lgbm        = error_estimator(lgbm,        observed, "lgbm")$rmse,
    .groups = "drop"
  ) %>%
  knitr::kable(digits = 2, caption = "Test-period RMSE by model and lead time")
```

```{r test-timeseries, fig.dim = c(12, 5)}
# Test-period time series for t+1
all_test_results %>%
  filter(lead == 1) %>%
  pivot_longer(cols = c(observed, persistence, gam_regime, lgbm),
               names_to = "series", values_to = "turb_ntu") %>%
  ggplot(aes(x = date, y = turb_ntu, color = series)) +
  geom_line(linewidth = 0.5, alpha = 0.85) +
  scale_color_manual(values = c(observed = "black", persistence = "grey50",
                                gam_regime = "#2166ac", lgbm = "#d73027")) +
  scale_y_log10() +
  theme_bw() +
  labs(title = "Test period (2024–2025): t+1 forecast vs. observed p95 NTU",
       x = NULL, y = "p95 NTU (log scale)", color = NULL)
```

---

# Variable Importance (LightGBM)

```{r lgbm-varimp, fig.dim = c(8, 5)}
# Show feature importance for t+1 model
lgbm_t1 <- lgb.load(
  here("data/derived/models/turbidity_forecast/lgbm_daily_lead_t1.txt")
)

lgb.importance(lgbm_t1, percentage = TRUE) %>%
  as_tibble() %>%
  slice_max(Gain, n = 15) %>%
  ggplot(aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  theme_bw() +
  labs(title = "LightGBM (t+1): Top 15 features by mean Gain",
       x = NULL, y = "Mean Gain (%)")
```

---

# Save GAM Objects

```{r save-gams}
saveRDS(final_gam_models,
        here("data/derived/models/turbidity_forecast/final_gam_models.rds"))
cat("GAM models saved.\n")
```
