---
title: "06_forecast_data_download — Daily p95 Turbidity Forecast (Modified Kemper)"
author: "ROSSyndicate"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(here)
library(tidyverse)
library(lubridate)
library(httr)
library(jsonlite)
library(arrow)

source(here("src/hefs/get_hefs_ensembles.R"))
source(here("src/hefs/get_hefs_gauges.R"))
```

# Purpose

Download HEFS (Hydro-Ensemble Forecast System) ensemble streamflow forecasts
for the Canyon Mouth gauge. HEFS serves as the operational discharge driver
for turbidity forecasting, analogous to the National Water Model medium-range
forecasts used by Kemper et al.

HEFS API: https://api.water.noaa.gov/hefs/v1/docs/

The Canyon Mouth gauge on the NWPS network is `COLOC070` (confirm in
`01_data_discovery.Rmd` if this has changed). HEFS provides 30-member ensemble
discharge forecasts at sub-daily timesteps; we aggregate to daily means.

**Coverage target:** 2023-01-01 to present (or most recent available date).
Earlier dates are not archived in the live HEFS API. For evaluation against
training data (2010–2023), see `09_evaluate_performance.Rmd` — that script
uses observed flows as a perfect-forecast benchmark.

---

# Constants

```{r constants}
# NWPS gauge ID for Canyon Mouth — confirm from 01_data_discovery.Rmd
CANYON_MOUTH_HEFS_ID <- "COLOC070"

# Download window
HEFS_START_DATE <- as.Date("2023-01-01")
HEFS_END_DATE   <- Sys.Date() - 1  # through yesterday

# Output path
OUT_PATH <- here("data/derived/models/turbidity_forecast/hefs_forecasts_raw.parquet")
```

---

# Verify Gauge

```{r verify-gauge}
hefs_gauges <- get_hefs_gauges()

hefs_gauges %>%
  filter(lid == CANYON_MOUTH_HEFS_ID) %>%
  select(lid, name, state, rfc) %>%
  print()
```

---

# Download HEFS Archive

We loop over each forecast initialization date from `HEFS_START_DATE` to
`HEFS_END_DATE`. HEFS updates once daily (around 12:00Z); we pull each day's
forecast. Failed dates are silently skipped and logged.

```{r download-hefs}
all_dates <- seq.Date(HEFS_START_DATE, HEFS_END_DATE, by = "day")
failed_dates <- character(0)

# Check if a partial archive already exists and only download new dates
if (file.exists(OUT_PATH)) {
  existing <- read_parquet(OUT_PATH)
  already_pulled <- unique(as.Date(existing$forecast_date))
  remaining_dates <- all_dates[!all_dates %in% already_pulled]
  cat("Existing archive covers", length(already_pulled), "dates.\n")
  cat("Downloading", length(remaining_dates), "new dates...\n")
} else {
  existing <- NULL
  remaining_dates <- all_dates
  cat("No existing archive. Downloading", length(remaining_dates), "dates...\n")
}

new_forecasts <- map(remaining_dates, function(d) {
  result <- tryCatch(
    get_hefs_ensemble(ID = CANYON_MOUTH_HEFS_ID, date = d),
    error = function(e) {
      failed_dates <<- c(failed_dates, as.character(d))
      NULL
    }
  )
  if (!is.null(result)) result %>% mutate(init_date = d)
  else NULL
}) %>%
  compact() %>%
  bind_rows()

if (nrow(new_forecasts) > 0) {
  cat("Downloaded", nrow(new_forecasts), "new forecast rows.\n")
} else {
  cat("No new forecast data downloaded.\n")
}

if (length(failed_dates) > 0) {
  cat("Failed dates:", paste(failed_dates, collapse = ", "), "\n")
}
```

---

# Aggregate to Daily

HEFS returns sub-daily (typically 6-hourly) ensemble members. Aggregate to
daily mean discharge per ensemble member per lead day.

```{r aggregate-daily}
if (nrow(new_forecasts) > 0) {
  hefs_daily_new <- new_forecasts %>%
    mutate(
      forecast_date = as.Date(date_time),
      lead_days     = as.integer(forecast_date - init_date)
    ) %>%
    filter(lead_days >= 1, lead_days <= 10) %>%  # keep up to 10-day lead
    group_by(init_date, forecast_date, lead_days, ensemble_member) %>%
    summarise(
      mean_q_cfs = mean(value * 35.3147, na.rm = TRUE),  # convert CMS → CFS
      .groups = "drop"
    )

  # Merge with existing archive
  hefs_daily <- bind_rows(existing, hefs_daily_new) %>%
    arrange(init_date, forecast_date, ensemble_member)
} else {
  hefs_daily <- existing
}

cat("Total HEFS archive rows (daily, all ensemble members):",
    nrow(hefs_daily), "\n")
cat("Date range (init_date):",
    as.character(range(hefs_daily$init_date)), "\n")
cat("Lead days covered:", paste(sort(unique(hefs_daily$lead_days)), collapse = ", "), "\n")
cat("Ensemble members:", length(unique(hefs_daily$ensemble_member)), "\n")
```

---

# Summarise Ensemble

Compute ensemble mean and quantiles (p10, p50, p90) per init_date × lead_days.
These are the primary features used in `07_forecast_data_prep.Rmd`.

```{r ensemble-summary}
hefs_summary <- hefs_daily %>%
  group_by(init_date, forecast_date, lead_days) %>%
  summarise(
    hefs_mean_cfs = mean(mean_q_cfs, na.rm = TRUE),
    hefs_p10_cfs  = quantile(mean_q_cfs, 0.10, na.rm = TRUE),
    hefs_p50_cfs  = quantile(mean_q_cfs, 0.50, na.rm = TRUE),
    hefs_p90_cfs  = quantile(mean_q_cfs, 0.90, na.rm = TRUE),
    hefs_spread   = hefs_p90_cfs - hefs_p10_cfs,
    n_members     = n_distinct(ensemble_member),
    .groups = "drop"
  )

cat("HEFS summary rows:", nrow(hefs_summary), "\n")
```

---

# Quick Sanity Plot

```{r hefs-plot, fig.dim = c(12, 4)}
# Compare HEFS ensemble mean to observed flow for a recent period
obs_flow <- cdssr::get_sw_ts(
  abbrev     = "CLAFTCCO",
  start_date = as.character(HEFS_START_DATE),
  end_date   = as.character(HEFS_END_DATE),
  timescale  = "day"
) %>%
  mutate(date = as.Date(datetime)) %>%
  select(date, observed_cfs = value)

hefs_summary %>%
  filter(lead_days == 1) %>%
  left_join(obs_flow, by = c("forecast_date" = "date")) %>%
  pivot_longer(cols = c(hefs_mean_cfs, observed_cfs),
               names_to = "series", values_to = "cfs") %>%
  ggplot(aes(x = forecast_date, y = cfs, color = series)) +
  geom_line(linewidth = 0.5, alpha = 0.8) +
  scale_color_manual(values = c(hefs_mean_cfs = "steelblue",
                                observed_cfs  = "black")) +
  scale_y_log10() +
  theme_bw() +
  labs(title = "HEFS 1-day-ahead ensemble mean vs. observed Canyon Mouth flow",
       x = NULL, y = "Discharge (CFS, log scale)", color = NULL)
```

---

# Save

```{r save}
write_parquet(hefs_daily,
              here("data/derived/models/turbidity_forecast/hefs_forecasts_raw.parquet"))
write_parquet(hefs_summary,
              here("data/derived/models/turbidity_forecast/hefs_forecasts_summary.parquet"))

cat("Saved raw ensemble:", here("data/derived/models/turbidity_forecast/hefs_forecasts_raw.parquet"), "\n")
cat("Saved summary:", here("data/derived/models/turbidity_forecast/hefs_forecasts_summary.parquet"), "\n")
```
