---
title: "07_forecast_data_prep — Daily p95 Turbidity Forecast (Modified Kemper)"
author: "ROSSyndicate"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(here)
library(tidyverse)
library(lubridate)
library(slider)
library(arrow)

source(knitr::purl(here("modeling/turbidity/modified_kemper_application_daily/00_functions.Rmd"),
                   quiet = TRUE))
```

# Purpose

Prepare the HEFS ensemble forecast data as predictor features for the
turbidity models. For each forecast initialization date, this script:

1. Uses **observed** flow, SWE, and precipitation up to and including the
   initialization date for antecedent features (lookback window).
2. Substitutes **HEFS ensemble mean flow** for the forecast date(s) as the
   current-day flow predictor.
3. Computes regime classification using observed conditions at initialization.
4. Outputs one row per `init_date × lead_days`, with all features needed
   for `08_make_forecasts.Rmd`.

Analogous to `kemper_turb_method/07_forecast_data_prep.Rmd` but adapted for
daily resolution and the addition of snow/melt features.

---

# Load Data

```{r load-data}
# Observed historical features (from 03_feature_engineering.Rmd)
model_data <- read_parquet(
  here("data/derived/models/turbidity_forecast/daily_model_data.parquet")
)

# HEFS ensemble summary (from 06_forecast_data_download.Rmd)
hefs_summary <- read_parquet(
  here("data/derived/models/turbidity_forecast/hefs_forecasts_summary.parquet")
)

cat("Model data rows:", nrow(model_data), "\n")
cat("HEFS summary rows:", nrow(hefs_summary), "\n")
cat("HEFS init_date range:", as.character(range(hefs_summary$init_date)), "\n")
```

---

# Build Antecedent Context at Each Initialization Date

For each init_date, we need the observed antecedent flow history to compute
rolling means, maxima, and time-since-peak features. We also need the current
(init_date) observed SWE, precip, and regime classification.

```{r antecedent-context}
# All unique initialization dates in the HEFS archive
init_dates <- sort(unique(hefs_summary$init_date))
cat("Initialisation dates:", length(init_dates), "\n")

# For each init_date, extract the row from model_data that corresponds to
# conditions at that date (antecedent features are already computed there)
init_context <- model_data %>%
  filter(date %in% init_dates) %>%
  select(
    date,
    # Current observed flow and state
    canyon_mouth_cfs, log_q, log_q_sq,
    # Antecedent flow features
    mean_q_7d, mean_q_30d, mean_q_90d,
    max_q_30d, max_q_90d, max_q_365d,
    time_since_30d_peak, time_since_90d_peak, time_since_annual_peak,
    log_limb,
    # Snow features
    swe_pc1, melt_rate_7d, melt_rate_14d,
    # Precipitation features
    precip, precip_3day, precip_7day, precip_event,
    # Temperature features (NA before 2018)
    tmax_c, tmax_5d_mean, swe_x_tmax,
    # Temporal
    doy, year,
    # Regime at init date
    regime,
    # Today's observed p95 (used for persistence baseline)
    p95
  ) %>%
  rename(init_date = date)
```

---

# Merge HEFS Flow with Init Context

For each lead day, replace `canyon_mouth_cfs` and `log_q` with the HEFS
ensemble mean forecast flow (the primary forecast driver), while keeping
all antecedent features from the initialization date.

The HEFS spread (`hefs_spread`) is retained as an additional predictor
capturing forecast uncertainty.

```{r merge-hefs-features}
hefs_features <- hefs_summary %>%
  filter(lead_days %in% 1:3) %>%
  left_join(init_context, by = "init_date") %>%
  mutate(
    # Override current-day flow features with HEFS-forecast values
    canyon_mouth_cfs_forecast = hefs_mean_cfs,
    log_q_forecast            = log1p(hefs_mean_cfs),
    log_q_sq_forecast         = log_q_forecast^2,

    # Approximate deltaQ: HEFS forecast minus last observed (at init_date)
    deltaQ_24h_forecast       = hefs_mean_cfs - canyon_mouth_cfs,
    deltaQ_48h_forecast       = case_when(
      lead_days == 1 ~ hefs_mean_cfs - canyon_mouth_cfs,
      lead_days >= 2 ~ hefs_mean_cfs -
        lag(hefs_mean_cfs, 1, default = NA_real_),
      .default = NA_real_
    ),

    # Log-limb: change in log-Q from init to forecast date
    log_limb_forecast = log_q_forecast - log_q,

    # dq_q: normalised flow change
    dq_q_forecast = (hefs_mean_cfs - canyon_mouth_cfs) / canyon_mouth_cfs
  ) %>%
  # Rename forecast-day flow features for clarity downstream
  rename(
    log_q_obs      = log_q,       # observed at init
    log_q          = log_q_forecast,  # HEFS-forecast (used as model input)
    log_q_sq       = log_q_sq_forecast,
    deltaQ_24h     = deltaQ_24h_forecast,
    deltaQ_48h     = deltaQ_48h_forecast,
    log_limb       = log_limb_forecast,
    dq_q           = dq_q_forecast
  )

cat("HEFS feature rows:", nrow(hefs_features), "\n")
cat("Lead days:", sort(unique(hefs_features$lead_days)), "\n")
```

---

# Verify Completeness

```{r completeness}
FORECAST_MODEL_COLS <- c("log_q", "deltaQ_24h", "melt_rate_7d",
                         "swe_pc1", "precip_3day", "precip_event",
                         "mean_q_7d", "mean_q_30d",
                         "doy", "regime")

hefs_features %>%
  summarise(across(all_of(FORECAST_MODEL_COLS),
                   ~ sum(is.na(.)), .names = "na_{.col}")) %>%
  pivot_longer(everything(), names_to = "col", values_to = "n_na") %>%
  filter(n_na > 0) %>%
  mutate(pct_na = round(100 * n_na / nrow(hefs_features), 1)) %>%
  arrange(desc(n_na))
```

---

# Spot-Check: HEFS vs. Observed Flow

```{r spot-check, fig.dim = c(12, 4)}
spot <- hefs_features %>%
  filter(lead_days == 1) %>%
  left_join(
    model_data %>%
      select(date, observed_cfs = canyon_mouth_cfs),
    by = c("forecast_date" = "date")
  )

spot %>%
  pivot_longer(cols = c(canyon_mouth_cfs_forecast, observed_cfs),
               names_to = "series", values_to = "cfs") %>%
  ggplot(aes(x = forecast_date, y = cfs, color = series)) +
  geom_line(linewidth = 0.5, alpha = 0.8) +
  scale_color_manual(
    values = c(canyon_mouth_cfs_forecast = "steelblue", observed_cfs = "black")
  ) +
  scale_y_log10() +
  theme_bw() +
  labs(title = "HEFS 1-day forecast vs. observed Canyon Mouth flow",
       x = NULL, y = "CFS (log scale)", color = NULL)
```

---

# Save

```{r save}
out_path <- here("data/derived/models/turbidity_forecast/hefs_forecast_features.parquet")
write_parquet(hefs_features, out_path)
cat("Saved:", out_path, "\n")
cat("Columns:", paste(names(hefs_features), collapse = ", "), "\n")
```
