---
title: "08_make_forecasts — Daily p95 Turbidity Forecast (Modified Kemper)"
author: "ROSSyndicate"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(here)
library(tidyverse)
library(lubridate)
library(arrow)
library(mgcv)
library(lightgbm)

source(knitr::purl(here("modeling/turbidity/modified_kemper_application_daily/00_functions.Rmd"),
                   quiet = TRUE))
```

# Purpose

Apply trained GAM and LightGBM models to HEFS forecast features to generate
operational 1–3 day p95 turbidity forecasts at the FC Intake.

Additionally, propagates HEFS ensemble uncertainty through the LightGBM model
(applying the model to each of the 30 ensemble members) to produce a forecast
distribution (p10, p50, p90).

Output: `turbidity_forecasts_v{date}.parquet`

---

# Load Models and Data

```{r load-models}
# Trained GAMs
final_gam_models <- readRDS(
  here("data/derived/models/turbidity_forecast/final_gam_models.rds")
)

# LightGBM metadata (smearing factors + predictor lists)
lgbm_model_meta <- readRDS(
  here("data/derived/models/turbidity_forecast/lgbm_model_meta.rds")
)

# Load LightGBM model objects
lgbm_models <- map(1:3, function(lead) {
  lgb.load(here("data/derived/models/turbidity_forecast",
                paste0("lgbm_daily_lead_t", lead, ".txt")))
})
names(lgbm_models) <- paste0("lead_t", 1:3)
```

```{r load-data}
# HEFS forecast features (from 07_forecast_data_prep.Rmd)
hefs_features <- read_parquet(
  here("data/derived/models/turbidity_forecast/hefs_forecast_features.parquet")
) %>%
  mutate(regime = factor(regime, levels = c("snowmelt", "convective", "baseflow")),
         regime_int = as.integer(regime))

# Raw ensemble members for uncertainty propagation
hefs_raw <- read_parquet(
  here("data/derived/models/turbidity_forecast/hefs_forecasts_raw.parquet")
)

# Observed model data for context (init_date antecedent features)
model_data <- read_parquet(
  here("data/derived/models/turbidity_forecast/daily_model_data.parquet")
)

cat("HEFS feature rows:", nrow(hefs_features), "\n")
cat("Init dates to forecast:", n_distinct(hefs_features$init_date), "\n")
```

---

# Generate Deterministic Forecasts

Apply GAM (global + regime) and LightGBM ensemble mean forecasts for leads
t+1, t+2, t+3.

```{r deterministic-forecasts}
det_forecasts <- map_dfr(1:3, function(lead) {
  lead_nm <- paste0("lead_t", lead)
  gam_mods <- final_gam_models[[lead_nm]]
  lgbm_mod  <- lgbm_models[[lead_nm]]
  meta      <- lgbm_model_meta[[lead_nm]]

  feat <- hefs_features %>% filter(lead_days == lead)

  # ── GAM predictions ──────────────────────────────────────────────────────────
  gam_global <- tryCatch(
    as.numeric(predict(gam_mods$global, newdata = feat, type = "response")),
    error = function(e) rep(NA_real_, nrow(feat))
  )

  gam_regime <- map2_dbl(
    as.character(feat$regime), seq_len(nrow(feat)),
    function(r, i) {
      tryCatch(
        as.numeric(predict(gam_mods$regime[[r]],
                           newdata = feat[i, ], type = "response")),
        error = function(e) NA_real_
      )
    }
  )

  # ── LightGBM predictions ─────────────────────────────────────────────────────
  # Rename regime to integer for LightGBM
  feat_lgbm <- feat %>%
    mutate(regime = regime_int)

  lgbm_pred <- tryCatch({
    test_mat  <- data.matrix(feat_lgbm %>% select(all_of(meta$preds)))
    log_pred  <- predict(lgbm_mod, data = test_mat)
    (10^log_pred) * meta$smear_d
  }, error = function(e) rep(NA_real_, nrow(feat)))

  tibble(
    init_date     = feat$init_date,
    forecast_date = feat$forecast_date,
    lead_days     = lead,
    regime        = feat$regime,
    hefs_mean_cfs = feat$hefs_mean_cfs,
    obs_p95_init  = feat$p95,          # today's observed (persistence baseline)
    gam_global    = gam_global,
    gam_regime    = gam_regime,
    lgbm_mean     = lgbm_pred
  )
})

cat("Deterministic forecast rows:", nrow(det_forecasts), "\n")
```

---

# Propagate HEFS Ensemble Uncertainty Through LightGBM

Apply LightGBM to each of the 30 HEFS ensemble members to obtain a forecast
distribution. Uses the same antecedent features from the init_date but swaps
in each member's forecast flow.

```{r ensemble-uncertainty}
# Build per-member feature tables for each init_date × lead
# Merge raw ensemble with init_date context from hefs_features (t+1 only for efficiency)

init_context <- hefs_features %>%
  select(init_date, lead_days,
         canyon_mouth_cfs,           # observed flow at init_date
         mean_q_7d, mean_q_30d, mean_q_90d,
         max_q_30d, max_q_90d, max_q_365d,
         time_since_30d_peak, time_since_90d_peak, time_since_annual_peak,
         swe_pc1, melt_rate_7d, melt_rate_14d,
         precip, precip_3day, precip_7day, precip_event,
         tmax_5d_mean, swe_x_tmax,
         doy, regime_int)

ens_forecasts <- map_dfr(1:3, function(lead) {
  meta     <- lgbm_model_meta[[paste0("lead_t", lead)]]
  lgbm_mod <- lgbm_models[[paste0("lead_t", lead)]]

  # Per-member daily flows at the forecast date
  member_flows <- hefs_raw %>%
    mutate(
      init_date     = as.Date(forecast_date) - lead,
      forecast_date = as.Date(date_time),
      lead_days     = lead
    ) %>%
    filter(lead_days == lead) %>%
    group_by(init_date, forecast_date, lead_days, ensemble_member) %>%
    summarise(member_q_cfs = mean(value * 35.3147, na.rm = TRUE),
              .groups = "drop")

  ens_feat <- member_flows %>%
    left_join(init_context %>% filter(lead_days == lead),
              by = c("init_date", "lead_days")) %>%
    drop_na(member_q_cfs, mean_q_7d) %>%
    mutate(
      log_q     = log1p(member_q_cfs),
      log_q_sq  = log_q^2,
      deltaQ_24h = member_q_cfs - canyon_mouth_cfs,
      log_limb   = log_q - log1p(canyon_mouth_cfs),
      dq_q       = (member_q_cfs - canyon_mouth_cfs) / canyon_mouth_cfs,
      deltaQ_48h = deltaQ_24h,   # approximation for members
      regime     = regime_int
    )

  # Apply model to each member
  ens_mat  <- data.matrix(ens_feat %>% select(all_of(meta$preds)))
  log_pred <- predict(lgbm_mod, data = ens_mat)
  ens_pred <- (10^log_pred) * meta$smear_d

  ens_feat %>%
    mutate(lead_days = lead,
           ens_pred_ntu = ens_pred) %>%
    group_by(init_date, forecast_date, lead_days) %>%
    summarise(
      lgbm_p10 = quantile(ens_pred_ntu, 0.10, na.rm = TRUE),
      lgbm_p50 = quantile(ens_pred_ntu, 0.50, na.rm = TRUE),
      lgbm_p90 = quantile(ens_pred_ntu, 0.90, na.rm = TRUE),
      n_members = n_distinct(ensemble_member),
      .groups = "drop"
    )
})

cat("Ensemble forecast rows:", nrow(ens_forecasts), "\n")
```

---

# Assemble Final Forecast Table

```{r assemble-forecasts}
turbidity_forecasts <- det_forecasts %>%
  left_join(ens_forecasts,
            by = c("init_date", "forecast_date", "lead_days")) %>%
  mutate(
    # Combine GAM and LightGBM into a simple ensemble mean
    ensemble_mean = rowMeans(cbind(gam_regime, lgbm_mean), na.rm = TRUE)
  ) %>%
  select(
    init_date, forecast_date, lead_days, regime,
    obs_p95_init,             # persistence baseline
    gam_global, gam_regime,   # GAM predictions
    lgbm_mean,                # LightGBM deterministic
    lgbm_p10, lgbm_p50, lgbm_p90,  # LightGBM ensemble uncertainty
    ensemble_mean,            # simple average of gam_regime + lgbm_mean
    hefs_mean_cfs,
    n_members
  )

cat("Final forecast rows:", nrow(turbidity_forecasts), "\n")
glimpse(turbidity_forecasts)
```

---

# Forecast Visualisation

```{r forecast-plot, fig.dim = c(12, 6)}
# Most recent init_date available
latest_init <- max(turbidity_forecasts$init_date)

turbidity_forecasts %>%
  filter(init_date == latest_init) %>%
  pivot_longer(cols = c(gam_regime, lgbm_mean, ensemble_mean),
               names_to = "model", values_to = "forecast_ntu") %>%
  ggplot(aes(x = forecast_date, y = forecast_ntu, color = model)) +
  geom_ribbon(
    data = turbidity_forecasts %>% filter(init_date == latest_init),
    aes(x = forecast_date, ymin = lgbm_p10, ymax = lgbm_p90),
    fill = "#d73027", alpha = 0.15, inherit.aes = FALSE
  ) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c(gam_regime = "#2166ac",
                                lgbm_mean  = "#d73027",
                                ensemble_mean = "black")) +
  scale_y_log10() +
  theme_bw() +
  labs(
    title = paste0("p95 Turbidity Forecast — initialized ", latest_init),
    subtitle = "Shaded band = LightGBM p10–p90 (HEFS ensemble spread)",
    x = "Forecast date", y = "Forecast p95 NTU (log scale)", color = "Model"
  )
```

---

# Save

```{r save}
out_dir <- here("data/derived/models/turbidity_forecast")
out_file <- file.path(out_dir, paste0("turbidity_forecasts_v", Sys.Date(), ".parquet"))
write_parquet(turbidity_forecasts, out_file)
cat("Saved:", out_file, "\n")
```
