---
title: "09_evaluate_performance — Daily p95 Turbidity Forecast (Modified Kemper)"
author: "ROSSyndicate"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(here)
library(tidyverse)
library(lubridate)
library(arrow)
library(hydroGOF)
library(patchwork)

source(knitr::purl(here("modeling/turbidity/modified_kemper_application_daily/00_functions.Rmd"),
                   quiet = TRUE))
```

# Purpose

Comprehensive evaluation of turbidity forecast skill using:
1. **Overall metrics** — RMSE, MAE, NSE, KGE (decomposed), PBIAS, r, R²
2. **Lead-time analysis** — skill decay from t+1 to t+3
3. **Regime-stratified metrics** — snowmelt vs. convective vs. baseflow
4. **High-turbidity event evaluation** — events > 10 NTU and > 75 NTU
5. **Event catalog comparison** — peak p95 error per event

Two evaluation periods are reported:
- **Cross-validation (2010–2023):** from `05_model_development.Rmd` CV predictions
- **Test period (2024–2025):** held-out independent evaluation

---

# Constants

```{r constants}
TURB_EVENT_THRESH <- 10   # NTU — elevated
TURB_HIGH_THRESH  <- 75   # NTU — high-turbidity treatment concern

MODEL_NAMES <- c("persistence", "gam_global", "gam_regime", "lgbm_mean", "ensemble_mean")
MODEL_COLORS <- c(
  persistence   = "grey50",
  gam_global    = "#7fbfff",
  gam_regime    = "#2166ac",
  lgbm_mean     = "#d73027",
  ensemble_mean = "black"
)
```

---

# Load Data

```{r load-data}
# Forecast output from 08_make_forecasts.Rmd
forecast_files <- list.files(
  here("data/derived/models/turbidity_forecast"),
  pattern = "^turbidity_forecasts_v.+\\.parquet$",
  full.names = TRUE
)
# Use most recent file
forecasts <- read_parquet(forecast_files[length(forecast_files)])

# Observed p95 for matching to forecast dates
observed <- read_parquet(
  here("data/derived/models/turbidity_forecast/daily_model_data.parquet")
) %>%
  select(date, p95_observed = p95, regime)

# Event catalog from 04_event_delineation.Rmd
event_catalog <- read_parquet(
  here("data/derived/models/turbidity_forecast/turbidity_event_catalog.parquet")
)

cat("Forecast rows:", nrow(forecasts), "\n")
cat("Init date range:", as.character(range(forecasts$init_date)), "\n")
```

```{r merge-observed}
eval_data <- forecasts %>%
  left_join(observed %>% rename(forecast_date = date),
            by = "forecast_date",
            suffix = c("_fc", "_obs")) %>%
  rename(
    observed   = p95_observed,
    regime_obs = regime_obs
  ) %>%
  drop_na(observed)

cat("Evaluation rows (with observed p95):", nrow(eval_data), "\n")
```

---

# 1. Overall Performance Metrics

```{r overall-metrics}
overall_metrics <- eval_data %>%
  drop_na(all_of(MODEL_NAMES)) %>%
  group_by(lead_days) %>%
  summarise(
    across(
      all_of(MODEL_NAMES),
      ~ error_estimator(.x, observed, "model")$rmse,
      .names = "{.col}_rmse"
    ),
    .groups = "drop"
  )

cat("Overall RMSE by lead time:\n")
print(overall_metrics)
```

```{r full-metrics-table}
full_metrics <- map_dfr(MODEL_NAMES, function(mod) {
  eval_data %>%
    filter(!is.na(.data[[mod]]), !is.na(observed)) %>%
    group_by(lead_days) %>%
    summarise(
      error_estimator(.data[[mod]], observed, mod),
      .groups = "drop"
    )
}) %>%
  arrange(lead_days, model)

full_metrics %>%
  select(lead_days, model, rmse, mae, nse, kge, pbias, corr) %>%
  knitr::kable(digits = 3, caption = "Full performance metrics by model and lead time")
```

---

# 2. Lead-Time Skill Decay

```{r lead-decay-plot, fig.dim = c(8, 5)}
full_metrics %>%
  filter(model != "persistence") %>%
  ggplot(aes(x = lead_days, y = rmse, color = model, group = model)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  geom_hline(
    data = full_metrics %>% filter(model == "persistence") %>%
      group_by(lead_days) %>% summarise(persistence_rmse = mean(rmse)),
    aes(yintercept = persistence_rmse),
    linetype = "dashed", color = "grey50"
  ) +
  scale_color_manual(values = MODEL_COLORS[MODEL_NAMES != "persistence"]) +
  scale_x_continuous(breaks = 1:3) +
  theme_bw() +
  labs(title = "Forecast RMSE by lead time",
       subtitle = "Dashed = persistence baseline",
       x = "Lead time (days)", y = "RMSE (NTU)", color = "Model")
```

```{r nse-kge-plot, fig.dim = c(10, 4)}
full_metrics %>%
  select(lead_days, model, nse, kge) %>%
  pivot_longer(cols = c(nse, kge), names_to = "metric") %>%
  ggplot(aes(x = lead_days, y = value, color = model, group = model)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey70") +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  facet_wrap(~ metric) +
  scale_color_manual(values = MODEL_COLORS) +
  scale_x_continuous(breaks = 1:3) +
  theme_bw() +
  labs(title = "NSE and KGE by lead time",
       x = "Lead time (days)", y = NULL, color = "Model")
```

---

# 3. Regime-Stratified Metrics

```{r regime-metrics}
regime_metrics <- map_dfr(MODEL_NAMES, function(mod) {
  eval_data %>%
    filter(!is.na(.data[[mod]]), !is.na(observed)) %>%
    group_by(lead_days, regime = regime_fc) %>%
    summarise(
      error_estimator(.data[[mod]], observed, mod),
      n = n(),
      .groups = "drop"
    )
}) %>%
  arrange(regime, lead_days, model)

regime_metrics %>%
  select(lead_days, regime, model, rmse, nse, kge, n) %>%
  filter(lead_days == 1) %>%
  knitr::kable(digits = 3, caption = "t+1 metrics by regime (all models)")
```

```{r regime-rmse-plot, fig.dim = c(10, 5)}
regime_metrics %>%
  filter(model %in% c("persistence", "gam_regime", "lgbm_mean")) %>%
  ggplot(aes(x = lead_days, y = rmse, color = model, group = model)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  facet_wrap(~ regime) +
  scale_color_manual(values = MODEL_COLORS) +
  scale_x_continuous(breaks = 1:3) +
  theme_bw() +
  labs(title = "RMSE by regime and lead time",
       x = "Lead time (days)", y = "RMSE (NTU)", color = "Model")
```

---

# 4. High-Turbidity Event Evaluation

```{r high-turb-metrics}
# Metrics computed only on days when observed p95 exceeds thresholds
thresholds <- c(event = TURB_EVENT_THRESH, high = TURB_HIGH_THRESH)

ht_metrics <- map_dfr(names(thresholds), function(thresh_nm) {
  thresh <- thresholds[thresh_nm]
  map_dfr(MODEL_NAMES, function(mod) {
    eval_data %>%
      filter(observed > thresh, !is.na(.data[[mod]])) %>%
      group_by(lead_days) %>%
      summarise(
        error_estimator(.data[[mod]], observed, mod),
        n        = n(),
        threshold = paste0(">", thresh, " NTU"),
        .groups = "drop"
      )
  })
})

ht_metrics %>%
  filter(lead_days == 1) %>%
  select(threshold, model, rmse, mae, nse, pbias, n) %>%
  knitr::kable(digits = 3, caption = "High-turbidity subset metrics (t+1)")
```

---

# 5. Event Catalog Evaluation

Compare peak p95 forecasts against event-catalog peaks.

```{r event-eval}
# For each event in the catalog, find the forecast issued closest to the event peak
event_eval <- event_catalog %>%
  mutate(peak_date = event_start + days(which.max(peak_p95))) %>%
  select(event_id, event_type, peak_date = event_start, peak_p95) %>%
  mutate(peak_date = event_start) %>%
  # Attempt to find a t+1 forecast for each event's start date
  left_join(
    eval_data %>% filter(lead_days == 1) %>%
      select(forecast_date, gam_regime, lgbm_mean, ensemble_mean),
    by = c("peak_date" = "forecast_date")
  ) %>%
  drop_na(gam_regime)

event_eval %>%
  mutate(
    gam_regime_err  = gam_regime  - peak_p95,
    lgbm_err        = lgbm_mean   - peak_p95,
    ensemble_err    = ensemble_mean - peak_p95
  ) %>%
  group_by(event_type) %>%
  summarise(
    n_events        = n(),
    mean_gam_err    = round(mean(gam_regime_err,  na.rm = TRUE), 1),
    mean_lgbm_err   = round(mean(lgbm_err,        na.rm = TRUE), 1),
    mean_ens_err    = round(mean(ensemble_err,     na.rm = TRUE), 1),
    rmse_gam        = round(sqrt(mean(gam_regime_err^2,   na.rm = TRUE)), 1),
    rmse_lgbm       = round(sqrt(mean(lgbm_err^2,         na.rm = TRUE)), 1),
    .groups = "drop"
  ) %>%
  knitr::kable(caption = "Event-based evaluation by event type (t+1 forecast at event start)")
```

```{r event-scatter, fig.dim = c(8, 4)}
event_eval %>%
  pivot_longer(cols = c(gam_regime, lgbm_mean),
               names_to = "model", values_to = "forecast") %>%
  ggplot(aes(x = peak_p95, y = forecast, color = event_type, shape = model)) +
  geom_point(size = 2.5, alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  scale_color_manual(
    values = c(snowmelt = "#2166ac", convective = "#d73027", "baseflow/other" = "#4dac26")
  ) +
  scale_x_log10() + scale_y_log10() +
  theme_bw() +
  labs(title = "Event peak: forecast vs. observed p95 NTU (t+1)",
       x = "Observed peak p95 (NTU)", y = "Forecast p95 (NTU)",
       color = "Event type", shape = "Model")
```

---

# 6. KGE Decomposition

```{r kge-decomposition}
kge_decomp <- map_dfr(MODEL_NAMES, function(mod) {
  eval_data %>%
    filter(!is.na(.data[[mod]]), !is.na(observed)) %>%
    group_by(lead_days) %>%
    summarise(
      model = mod,
      decompose_kge(.data[[mod]], observed),
      .groups = "drop"
    )
})

kge_decomp %>%
  filter(lead_days == 1) %>%
  select(model, r, bias, variability, kge_manual) %>%
  knitr::kable(digits = 3, caption = "KGE decomposition (t+1 forecasts)")
```

---

# 7. Forecast Time Series — Test Period

```{r test-ts-t1, fig.dim = c(14, 5)}
eval_data %>%
  filter(lead_days == 1) %>%
  pivot_longer(cols = c(observed, gam_regime, lgbm_mean, persistence),
               names_to = "series", values_to = "ntu") %>%
  ggplot(aes(x = forecast_date, y = ntu, color = series)) +
  geom_line(linewidth = 0.5, alpha = 0.85) +
  geom_ribbon(
    data = eval_data %>% filter(lead_days == 1,
                                !is.na(lgbm_p10), !is.na(lgbm_p90)),
    aes(x = forecast_date, ymin = lgbm_p10, ymax = lgbm_p90),
    fill = "#d73027", alpha = 0.10, inherit.aes = FALSE
  ) +
  scale_color_manual(values = c(observed    = "black",
                                gam_regime  = "#2166ac",
                                lgbm_mean   = "#d73027",
                                persistence = "grey50")) +
  scale_y_log10() +
  theme_bw() +
  labs(title = "t+1 forecast vs. observed p95 NTU — full evaluation period",
       subtitle = "Shaded band = LightGBM ensemble p10–p90",
       x = NULL, y = "p95 NTU (log scale)", color = NULL)
```

---

# Summary Table

```{r summary-table}
full_metrics %>%
  select(lead_days, model, rmse, nse, kge, pbias) %>%
  pivot_longer(cols = c(rmse, nse, kge, pbias)) %>%
  pivot_wider(names_from = c(name, lead_days),
              names_glue = "{name}_t{lead_days}") %>%
  knitr::kable(digits = 3, caption = "Summary: all metrics across lead times")
```
