---
title: "overview"
author: "Sam Struthers"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
#load all packages and themes
source("src/setup_libraries.R")

# Configure parallel processing
max_workers <- 4  
# Maximum number of parallel workers# set up parallel processing
num_workers <- min(availableCores() - 1, max_workers)
plan(multisession, workers = num_workers)
furrr_options(
  globals = TRUE,
  packages = c("arrow", "data.table", "httr2", "tidyverse", "lubridate", "zoo",
               "padr", "stats", "RcppRoll", "yaml", "here", "fcw.qaqc")
)

# suppress scientific notation for consistent formatting
options(scipen = 999)
```

# Pull in cached data from some previous GH actions run


```{r}
cached_data

max_date_site 
```


# Setting up dates

TODO: redo so that it is only pulling the data needed from most recent GH pull

```{r}

start_DT <- ymd_hms("2025-04-01 00:00:00", tz = "America/Denver")
#start_DT <- Sys.time() - days(7) # Set the start date to 7 days ago
end_DT <- Sys.time() # Set the end date to now

```

# Pulling in data

A few notes: When present, the column `DT_round` will be in UTC and `DT_round_MT` will be in Mountain Time. The column `value` is the raw value recorded at that time. The column `parameter` is the name of the parameter being measured (e.g., "water_temperature", "turbidity", etc.). The column `site` is the acronym of the site where the data was collected.

## Radio Telemetry Data

This comes through the WET API service which creates a web page for each site and sensor. This is pulled in by the function `pull_wet_api()`. In this chunk we are loading in all the data for the three sites (sfm, chd, pfal) for the last week. 

```{r}
# these values were used in testing or are invalid values that we want to filter out
#TODO: We may not want to filter them at this step but will later on so that I can see when a sensor is down
invalid_values <- c(-9999, 638.30, -99.99)

source("src/pull_wet_api.R")
# Sites livestreaming via this method
sites <- c("sfm", "chd", "pfal")

wet_data <- map(sites,
                ~pull_wet_api(
                  target_site = .x,
                  start_datetime = start_DT,
                  end_datetime = end_DT,
                  data_type = "all",
                  time_window = "all"
                ))%>%
  rbindlist()%>%
  #remove invalid values or NAs
  filter(value %nin% invalid_values, !is.na(value))%>%
  split(f = list(.$site, .$parameter), sep = "-")

```

## HydroVu Livestream Data

All of the lower sites livestream at a 3 hour interval but our Canyon Mouth site (located just above Greeley intake) does send to this service at a 1 hour interval (we can increase this if needed)


```{r}
# Set up parallel processing
num_workers <- min(availableCores() - 1, 4) # Use at most 4 workers
plan(multisession, workers = num_workers)
furrr_options(
  globals = TRUE,
  packages = c("arrow", "data.table", "httr2", "tidyverse", "lubridate", "zoo",
               "padr", "stats", "RcppRoll", "yaml", "here", "fcw.qaqc")
)

# suppress scientific notation to ensure consistent formatting
options(scipen = 999)

# Establishing staging directory - Replacing with temp_dir()
#staging_directory <- here("data", "api_pull", "raw")
staging_directory = tempdir()

# Read in credentials
hv_creds <- read_yaml(here("data", "upper_clp_dss", "creds", "HydroVuCreds.yml"))
hv_token <- hv_auth(client_id = as.character(hv_creds["client"]),
                    client_secret = as.character(hv_creds["secret"]))

# Pulling in the data from hydrovu
# Making the list of sites that we need
hv_sites <- hv_locations_all(hv_token) %>%
  filter(!grepl("vulink", name, ignore.case = TRUE))%>%
  #sondes with 2024 in the name can be avoided to speed up the live data pull 
  #these should be included in the historical data pull
  filter(!grepl("2024", name, ignore.case = TRUE))

# these sites are backed up on HydroVu but most do not livestream
# TODO: set up a daily CRON job that will look to see if any new data is available on HydroVu and grab it when possible?
# sites <- c( "cbri", "chd","joei",  "pbd", "pbr","pfal", "pman", "sfm") 

#Canyon mouth does send data to livestream
sites <- c("pbd")

# When we are getting all the 2025 data across the network for modeling, use these sites
#sites <- c("pbd", "salyer", "udall", "riverbend", "cottonwood", "springcreek" , "elc", "boxcreek",  "archery", "riverbluffs")
source(file = "src/api_puller.R")

walk(sites,
     function(site) {
       message("Requesting HV data for: ", site)
       api_puller(
         site = site,
         network = "all",
         start_dt = with_tz(start_DT, tzone = "UTC"), # api puller needs UTC dates
         end_dt = with_tz(end_DT, tzone = "UTC"),
         api_token = hv_token,
         hv_sites_arg = hv_sites,
         dump_dir = staging_directory
       )
     }
)

# read in data from staging directory

hv_data <- list.files(staging_directory, full.names = TRUE, pattern = ".parquet") %>%
  map_dfr(function(file_path){
    site_df <- read_parquet(file_path, as_data_frame = TRUE)
    return(site_df)
  }) %>%
  #doing some clean up
  select(-id) %>%
  mutate(units = as.character(units)) %>%
  #double check that Vulink data has been removed
  filter(!grepl("vulink", name, ignore.case = TRUE)) %>%
  mutate(
    DT = timestamp, #timestamp comes in UTC
    DT_round = round_date(DT, "15 minutes"), #rounding
    #DT_round_MT = with_tz(DT_round, tzone = "America/Denver"),
    DT_join = as.character(DT_round), #keeping in UTC but character form
    site = tolower(site)) %>%
  select(-name) %>%
  distinct(.keep_all = TRUE)%>%
  split(f = list(.$site, .$parameter), sep = "-") %>%
  keep(~nrow(.) > 0)

```


## Contrail Data
This is where Fort Collins' data livestreams. Since we have sites colocated there, we will define them as fc
Need to figure out if there is a more formal API for contrail but for now this will work...

```{r}

source("src/pull_contrail_api.R")
# Read/set up credentials
creds <- yaml::read_yaml(here("data","upper_clp_dss", "creds","contrail_creds.yml")) %>%
  unlist()
username <- as.character(creds["username"])
password <- as.character(creds["password"])
login_url <- as.character(creds["login_url"])

# Call the downloader function
contrail_data <- pull_contrail_api(start_DT, end_DT, username,password, login_url)

```



## Non livestreaming site data: Ignore for now!


These data are saved in the `sensor_data` folder and come in both VuLink and Aquatroll logs. All Aquatroll logs are uploaded to HydroVu but Vulink logs cannot be so we will read in both and then join any data not contained in the aquatroll log but that does exist in the VuLink log as a backup. 


```{r}
# 
# source("src/parse_insitu_html_log.R")
# #sites that do not livestream
# sites <- c( "cbri","joei", "pbr", "pman", "pfal") 
# 
# troll_files <- list.files(path = "data/sensor_data/2025/", pattern = "troll", full.names = T)%>%
#   #only look for our upper sites that do not livestream in any capacity
#   str_subset(paste(sites, collapse = "|"))
# 
# troll_data <- map(troll_files, parse_insitu_html_log) %>%
#   bind_rows()
# 
# vulink_files <- list.files(path = "data/sensor_data/2025/", pattern = "vulink", full.names = T)%>%
#   #only look for our upper sites that do not livestream in any capacity
#   str_subset(paste(sites, collapse = "|"))
# 
# vulink_data <- map(vulink_files, parse_insitu_html_log) %>%
#   bind_rows()
# 
# #always start with troll data and then fill in where needed with vulink data
# log_data <- troll_data %>%
#   #find the data that is not in the troll data but is in the vulink log and bind in
#   bind_rows(anti_join(vulink_data, ., by = c("site", "parameter", "DT_join")))%>%
#   #check for duplicates
#    distinct(.keep_all = TRUE) %>%
# # #format to match other data
#   split(f = list(.$site, .$parameter), sep = "-") %>%
#    keep(~nrow(.) > 0)


```

# Pull dataset

```{r}

write_rds(wet_data, "data/upper_clp_dss/api_pull/raw/wet_data_20250401_20250904.rds")
write_rds(hv_data, "data/upper_clp_dss/api_pull/raw/hv_data_20250401_20250904.rds")
write_rds(contrail_data, "data/upper_clp_dss/api_pull/raw/contrail_data_20250401_20250904.rds")



```



# Summarizing/joining Datasets

```{r}
# combine all data and remove duplicate site/sensor combos (from log data + livestream)
all_data_raw <- c(hv_data, wet_data, contrail_data)%>%
                  #, log_data )%>%
  bind_rows()%>%
  #convert depth to m for standardization with seasonal thresholds
  mutate(value = ifelse(parameter == "Depth" & units == "ft", value * 0.3048, value),
         units = case_when(parameter == "Depth" & units == "ft" ~ "m", 
                           parameter == "Temperature" & units == "C" ~ "°C",
                           TRUE ~ units), 
         timestamp = DT)%>%
  #TODO: remove this for backup?
  filter(between(DT_round, with_tz(start_DT, tzone = "UTC"), with_tz(end_DT, tzone = "UTC")))%>%
  split(f = list(.$site, .$parameter), sep = "-") %>%
  keep(~nrow(.) > 0)

# remove stage data 
list_names <- names(all_data_raw)
keep_indices <- !grepl("stage", list_names, ignore.case = TRUE)
all_data_raw <- all_data_raw[keep_indices]

# Tidy all the raw files
tidy_data <- all_data_raw %>%
  map(~tidy_api_data(api_data = .)) %>%  # the summarize interval default is 15 minutes
  keep(~!is.null(.))

```


## Saving tidy testing dataset

Can probably skip for GH actions version unless we want to commit multiple files

```{r}
write_rds(tidy_data, paste0("dashboard/data/tidy_data_subset_",as.Date(start_DT),"_",as.Date(end_DT),".rds"))

```



# Applying QAQC Code

## Reading in summarized dataset

```{r}
# Read in summarized data for dashboard
tidy_data <- read_rds(file = "dashboard/data/tidy_data_subset_2025-05-01_2025-08-29.rds")

#remove ORP data
tidy_data <- tidy_data %>%
  keep_at(imap_lgl(., ~ !grepl("ORP", .y)))
```


## Read in thresholds and sensor notes

These threshold files will likely need to be uploaded to GH for actions to work or will need to be in the secrets in GH

```{r}
# Configure your threshold files
sensor_thresholds_file <- "data/field_notes/qaqc/sensor_spec_thresholds.yml"
seasonal_thresholds_file <- "data/field_notes/qaqc/updated_seasonal_thresholds_2025_sjs.csv"
fc_seasonal_thresholds_file <- "data/field_notes/qaqc/fc_seasonal_thresholds_2025_sjs.csv"
fc_field_notes_file <- "data/sensor_data/FC_sondes/fc_field_notes_formatted.rds"
# read threshold data
sensor_thresholds <- read_yaml(sensor_thresholds_file)
fc_seasonal_thresholds <- read_csv(fc_seasonal_thresholds_file, show_col_types = FALSE)
#load all thresholds and bind with FC thresholds
season_thresholds <- read_csv(seasonal_thresholds_file, show_col_types = FALSE)%>%
  bind_rows(fc_seasonal_thresholds)

# Pulling in the data from mWater (where we record our field notes)
mWater_creds <- read_yaml(here("creds", "mWaterCreds.yml"))
mWater_data <- load_mWater(creds = mWater_creds)

#Summarized from provided notes in `data/sensor_data/FC_sondes/`
fc_field_notes <- read_rds(fc_field_notes_file)
#Joining field notes
all_field_notes <- grab_mWater_sensor_notes(mWater_api_data = mWater_data) %>%
  bind_rows(fc_field_notes)

# Grab sensor malfunction notes from mWater (We don't have records of malfunctions from FC so we will use ours as a placeholder)
sensor_malfunction_notes <- grab_mWater_malfunction_notes(mWater_api_data = mWater_data)%>% 
  #notes come in as MST, converting to UTC
  mutate(start_DT = with_tz(start_DT, tzone = "UTC"),
         end_DT = with_tz(end_DT, tzone = "UTC"))
```

## Adding field notes and summary statistics

```{r}
# Add the field note data to all of the data

combined_data <- tidy_data %>%
  future_map(~add_field_notes(df = ., notes = all_field_notes), .progress = TRUE)

# # Add summary statistics
summarized_data <- combined_data %>%
  map(~generate_summary_statistics(.))

#rm(combined_data, tidy_data)
```

## Single Sensor Flags


```{r}
# process data in chunks for memory efficiency
summarized_data_chunks <- split(1:length(summarized_data),
                                ceiling(seq_along(1:length(summarized_data))/10))

single_sensor_flags <- list()

for (chunk_idx in seq_along(summarized_data_chunks)) {
  message("\n=== Processing chunk ", chunk_idx, " of ", length(summarized_data_chunks), " ===")
  
  indices <- summarized_data_chunks[[chunk_idx]]
  chunk_data <- summarized_data[indices]
  
  # apply single-parameter flags
  chunk_results <- chunk_data %>%
    map(
      function(data) {
        flagged_data <- data %>%
          data.table(.) %>%
          # flag field visits
          add_field_flag(df = .) %>%
          # flag missing/NA values
          add_na_flag(df = .) %>%
          # flag dissolved oxygen noise patterns
          find_do_noise(df = .) %>%
          # # flag repeating/stuck values
          # add_repeat_flag(df = .) %>%
          # # flag depth shifts (sonde movement)
          # add_depth_shift_flag(df = ., level_shift_table = all_field_notes, post2024 = TRUE) %>%
          # flag sensor drift (FDOM, Chl-a, Turbidity)
          add_drift_flag(df = .)
        
        # apply sensor specification flags if thresholds exist
        if (unique(data$parameter) %in% names(sensor_thresholds)) {
          flagged_data <- flagged_data %>%
            data.table(.) %>%
            add_spec_flag(df = ., spec_table = sensor_thresholds)
        }
        
        # apply seasonal threshold flags if available
        if (unique(data$parameter) %in% unique(season_thresholds$parameter)) {
          flagged_data <- flagged_data %>%
            data.table(.) %>%
            add_seasonal_flag(df = ., threshold_table = season_thresholds)
        }
        
        return(flagged_data)
      },
      .progress = TRUE
    )
  
  single_sensor_flags <- c(single_sensor_flags, chunk_results)
  
  if (chunk_idx < length(summarized_data_chunks)) {
    gc()  # garbage collection between chunks
    Sys.sleep(0.1)
  }
}
```

## Intrasensor flags

```{r}
# combine single-parameter flags by site
intrasensor_flags <- single_sensor_flags %>%
  rbindlist(fill = TRUE) %>%
  split(by = "site")

# process inter-parameter flags in chunks
intrasensor_data_chunks <- split(1:length(intrasensor_flags),
                                 ceiling(seq_along(1:length(intrasensor_flags))/2))

intrasensor_flags_list <- list()
for (chunk_idx in seq_along(intrasensor_data_chunks)) {
  message("\n=== Processing chunk ", chunk_idx, " of ", length(intrasensor_data_chunks), " ===")
  
  indices <- intrasensor_data_chunks[[chunk_idx]]
  chunk_data <- intrasensor_flags[indices]
  
  chunk_results <- chunk_data %>%
    map(
      function(data) {
        flagged_data <- data %>%
          data.table() %>%
          # flag when water temperature below freezing
          add_frozen_flag(.) %>%
          # check for overlapping flags and resolve
          intersensor_check(.) %>%
          # flag potential sensor burial
          add_burial_flag(.) %>%
          # flag when sonde is above water surface
          add_unsubmerged_flag(.)
        
        return(flagged_data)
      }
    ) %>%
    rbindlist(fill = TRUE) %>%
    mutate(flag = ifelse(flag == "", NA, flag)) %>%
    split(f = list(.$site, .$parameter), sep = "-") %>%
     purrr::discard(~ nrow(.) == 0)%>%
    # # add known sensor malfunction periods
     map(~add_malfunction_flag(df = ., malfunction_records = sensor_malfunction_notes))
  
  intrasensor_flags_list <- c(intrasensor_flags_list, chunk_results)
  
  if (chunk_idx < length(intrasensor_data_chunks)) {
    gc()
    Sys.sleep(0.1)
  }
}
```

## Network Check and Final Flags

```{r}
#custom network check while fcw.qaqc package is being updated
source("src/network_check.R")

# apply network-level quality control
network_flags <- intrasensor_flags_list %>%
  # network check compares patterns across sites
  purrr::map(~network_check(df = .,network = "uclp_dashboard", intrasensor_flags_arg = intrasensor_flags_list)) %>%
  rbindlist(fill = TRUE) %>%
  # clean up flag column formatting
  tidy_flag_column() %>%
  split(f = list(.$site, .$parameter), sep = "-") %>%
  # add suspect data flags for isolated anomalies
  purrr::map(~add_suspect_flag(.)) %>%
  rbindlist(fill = TRUE)


# final data cleaning and preparation
v_final_flags <- network_flags %>%
  # Remove isolated suspect flags (single point anomalies)
  dplyr::mutate(auto_flag = ifelse(
    is.na(auto_flag), NA,
    ifelse(auto_flag == "suspect data" &
             is.na(lag(auto_flag, 1)) &
             is.na(lead(auto_flag, 1)), NA, auto_flag)
  )) %>%
  # select final columns
  dplyr::select(c("DT_round", "DT_join", "site", "parameter", "mean", "units",
                  "n_obs", "spread", "auto_flag", "mal_flag", "sonde_moved",
                  "sonde_employed", "season", "last_site_visit")) %>%
  # clean up empty flags
  dplyr::mutate(auto_flag = ifelse(is.na(auto_flag), NA,
                                   ifelse(auto_flag == "", NA, auto_flag))) %>%
  # split back into site-parameter combinations
  split(f = list(.$site, .$parameter), sep = "-") %>%
  keep(~nrow(.) > 0)


```

```{r}
ggplotly(v_final_flags%>%
    filter(parameter == "Temperature"& site %in% c( "pman_fc"))%>%
    ggplot(aes(DT_round, mean, color = auto_flag))+
    geom_point()+
      #scale_color_viridis_c()+
  #geom_hline(yintercept = 7.2, linetype = "dashed", color = "blue")+
   # geom_hline(yintercept = 8.8, linetype = "dashed", color = "blue")+
    facet_wrap(~site))#+
    #ylim(6,9))
```


# Write to new file

```{r}

# save as RDS file
write_rds(v_final_flags, here("dashboard", "data", "all_sensor_subset_flagged_2025-06-22_2025-08-08.rds"))
# Commit to repo/cache for later use

```
# End of GH actions run



# Sam Scratch

```{r}

plot_data <- v_final_flags%>%
  keep_at(imap_lgl(., ~ grepl("DO", .y))) %>%
  rbindlist()%>%
  mutate(DT_round = as.POSIXct(DT_round, tz = "America/Denver"))%>%
  filter(is.na(mal_flag))#%>%
  #remove flags that are only outside of seasonal range
  mutate(auto_flag = case_when(auto_flag == "outside of seasonal range" ~ NA,
                               auto_flag == "drift" ~ NA,
                               auto_flag == "drift;missing data" ~ NA,
                               TRUE ~ auto_flag))


ggplotly(ggplot(plot_data%>%filter(site %in% c( "pfal", "pbd") & parameter == "DO"), aes(x = DT_round, y = mean, color = auto_flag)) +
  geom_point() +
  facet_wrap(~site)+ #scales = "free_y") +
  #scale_y_log10()+
  labs(title = "Flagged Data",
       x = "Date",
       y = "Mean") +
  theme_minimal() +
  theme(legend.position = "bottom"))
```


# Create acceptable flags to show on dashboard

```{r}

flag_matrix <- tibble(
  parameter = c("Temperature", "pH", "Dissolved Oxygen", "Chl-a Fluorescence", "Turbidity", "FDOM Fluorescence", "Specific Conductivity"), 
  overflagging = c(NA,NA, NA, NA, NA, "outside of seasonal range|drift|drift;missing data", "outside of seasonal range" )
)


```



