---
title: "01_data_prep — FC Intake Turbidity (Kemper Method)"
author: "ROSSyndicate"
date: "`r Sys.Date()`"
output: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

This script loads and prepares the Fort Collins Intake turbidity and Canyon
Mouth discharge data for application of the Kemper et al. (2025) turbidity
forecasting method.

Adaptations from the original Kemper workflow:
- Turbidity comes from a local SCADA CSV (15-min, FNU) rather than USGS NWIS.
- Discharge comes from CDSSR `get_telemetry_ts()` at hourly resolution, enabling
the flow-weighted turbidity aggregation used in Kemper et al.
- There is a single flow station (no upstream tributary network).
- No probe-to-probe correction is needed (single sensor type).
- No tunnel-flow filter is applied.

Output: `intake_turbidity_plus_drivers_clean` — hourly modeling dataset with
flow-weighted turbidity and antecedent discharge features.


### Packages

```{r packages}
library(here)
library(tidyverse)
library(lubridate)
library(tsibble)
library(zoo)
library(slider)
library(cdssr)
library(arrow)
```

# Load Data

## Turbidity — FC Intake (15-min SCADA CSV)
```{r load-turbidity}
#### The raw file has a wide layout with many empty columns.
#### Only the first two columns are used:
####   col 1 — SampleDatetime  (format: YYYYMMdd HHMM, trailing space)
####   col 2 — N311            (turbidity, FNU)

raw_turb <- read_csv(
  here("data/raw/fc_intake_data/Intake_Turbidity_2010-2025.csv"),
  col_names = TRUE,
  col_select = 1:2,
  col_types = cols(.default = col_character()),
  show_col_types = FALSE
) %>%
  rename(raw_datetime = 1,
         turbidity_fnu = 2) %>%
  mutate(
    raw_datetime = str_trim(raw_datetime),
    dateTime = parse_date_time(raw_datetime,
                               orders = "Ymd HM",
                               tz = "America/Denver"),
    turbidity_fnu = as.numeric(turbidity_fnu)
  ) %>%
  filter(!is.na(dateTime),
         !is.na(turbidity_fnu),
         turbidity_fnu >= 0) %>%
  select(dateTime, turbidity_fnu)

glimpse(raw_turb)
```

## Discharge — Canyon Mouth (CLAFTCCO, CDSSR, hourly telemetry)

```{r load-flow}
#### get_telemetry_ts with timescale = "hour" returns observed hourly discharge
#### directly from the CDSSR telemetry network, eliminating the need for
#### interpolation from daily values.
#### Returns columns including `datetime` and `value` (cfs).
#### Gaps of ≤ 24 h are filled by linear interpolation.

flow_raw <- get_telemetry_ts(
  abbrev = "CLAFTCCO",
  parameter = "DISCHRG",
  start_date = "2010-01-01",
  end_date = "2025-12-31",
  timescale = "hour"
)

## TO-DO: need to deal with flags here

flow_hourly <- flow_raw %>%
  select(dateTime = datetime, flow_cfs = meas_value) %>%
  mutate(dateTime = as.POSIXct(dateTime, tz = "America/Denver")) %>%
  arrange(dateTime) %>%
  as_tsibble(index = "dateTime") %>%
  fill_gaps(.full = FALSE) %>%
  mutate(
    flow_cfs = na.approx(flow_cfs, maxgap = 24, na.rm = FALSE),
    flow_cms = pmax(flow_cfs * 0.0283168, 0.001) # cfs → cms; guard against zero
  ) %>%
  as_tibble() %>%
  mutate(
    log_observed_flow = log10(flow_cms),
    log_squared_observed_flow = log_observed_flow^2
  ) %>%
  filter(!is.na(flow_cms))

glimpse(flow_hourly)
```

# Turbidity QC

## Turbidity — Munroe (upstream reference, 15-min SCADA CSV)

```{r load-munroe}
#### Same wide-CSV layout as the intake file; only cols 1–2 are used.
#### Column 2 (N4118) is turbidity in FNU at the Munroe site, which is
#### slightly upstream of the FC Intake sensor.

munroe_raw <- read_csv(
  here("data/raw/fc_intake_data/Munroe_Turbidity_2013-2025.csv"),
  col_names = TRUE,
  col_select = 1:2,
  col_types = cols(.default = col_character()),
  show_col_types = FALSE
) %>%
  rename(raw_datetime = 1,
         munroe_fnu   = 2) %>%
  mutate(
    raw_datetime = str_trim(raw_datetime),
    dateTime     = parse_date_time(raw_datetime,
                                   orders = "Ymd HM",
                                   tz     = "America/Denver"),
    munroe_fnu   = as.numeric(munroe_fnu)
  ) %>%
  filter(!is.na(dateTime), !is.na(munroe_fnu), munroe_fnu >= 0) %>%
  select(dateTime, munroe_fnu)

glimpse(munroe_raw)
```

```{r munroe-lag-plot, fig.width = 12, fig.height = 5}
#### First-pass outlier check: aggregate both sensors to hourly means,
#### then lag Munroe by 1 h (travel time from Munroe → Intake).
#### Points where the intake sensor departs strongly from the Munroe
#### signal shifted forward 1 h are candidate outliers.

intake_hourly_raw <- raw_turb %>%
  mutate(hour_dt = floor_date(dateTime, "hour")) %>%
  summarise(intake_fnu = mean(turbidity_fnu, na.rm = TRUE),
            .by = hour_dt)

munroe_hourly <- munroe_raw %>%
  mutate(hour_dt = floor_date(dateTime, "hour")) %>%
  summarise(munroe_fnu = mean(munroe_fnu, na.rm = TRUE),
            .by = hour_dt) %>%
  mutate(hour_dt_lagged = hour_dt - hours(1))   # shift upstream → downstream

lag_joined <- intake_hourly_raw %>%
  inner_join(munroe_hourly %>% select(hour_dt_lagged, munroe_fnu),
             by = c("hour_dt" = "hour_dt_lagged"))

#### Time-series overlay
lag_joined %>%
  pivot_longer(c(intake_fnu, munroe_fnu),
               names_to = "sensor", values_to = "fnu") %>%
  mutate(sensor = recode(sensor,
                         intake_fnu = "FC Intake",
                         munroe_fnu = "Munroe (lag +1 h)")) %>%
  ggplot(aes(x = hour_dt, y = fnu, color = sensor)) +
  geom_line(linewidth = 0.25, alpha = 0.7) +
  scale_y_log10() +
  scale_color_manual(values = c("FC Intake" = "steelblue",
                                "Munroe (lag +1 h)" = "darkorange")) +
  labs(title = "Cross-sensor comparison — FC Intake vs. Munroe (1-h lag, log scale)",
       subtitle = "Divergences suggest candidate outliers in the intake record",
       x = NULL, y = "FNU", color = NULL) +
  theme_bw() +
  theme(legend.position = "top")

#### Scatter: Munroe (lag +1 h) vs. Intake — points far from the 1:1 line are suspects
lag_joined %>%
  ggplot(aes(x = munroe_fnu, y = intake_fnu)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(alpha = 0.15, size = 0.7, color = "steelblue") +
  scale_x_log10() +
  scale_y_log10() +
  labs(title = "Scatter: Munroe (lag +1 h) vs. FC Intake turbidity",
       subtitle = "Dashed line = 1:1; outliers appear above or below",
       x = "Munroe FNU (lagged 1 h)", y = "FC Intake FNU") +
  theme_bw()
```

---

```{r turb-qc-params}
#### Tunable QC thresholds — adjust after reviewing diagnostics below.
####
####   STUCK_MIN_RUN      — flag runs of ≥ N consecutive identical values
####                         (sensor offline returning last-known value)
####   SPIKE_ROC_THRESHOLD — flag if |Δturb| / prior value exceeds this ratio
####                         (e.g. 10 = 1000% change in a single 15-min step)

STUCK_MIN_RUN       <- 10
SPIKE_ROC_THRESHOLD <- 10
```

```{r turb-qc-diagnostic, fig.width = 12, fig.height = 4}
#### Full raw record — use this to spot gross anomalies before flagging.
map(year(raw_turb$dateTime), 
    ~ raw_turb %>%
      filter(year(dateTime) == .x) %>% 
      ggplot(aes(x = dateTime, y = turbidity_fnu)) +
      geom_line(linewidth = 0.2, alpha = 0.5, color = "steelblue") +
      scale_y_log10() +
      labs(title = "Raw 15-min Turbidity — full record (log scale)",
           x = NULL, y = "FNU") +
      theme_bw()
)
```

```{r turb-qc-flag}
#### Flag 1 — Stuck sensor: ≥ STUCK_MIN_RUN consecutive identical values.
#### This is the SCADA analog of Kemper's provisional-data filter:
#### sensors that drop offline typically latch onto their last valid reading.

raw_turb_flagged <- raw_turb %>%
  arrange(dateTime) %>%
  mutate(
    rl_id  = with(rle(turbidity_fnu), rep(seq_along(lengths), lengths)),
    rl_len = with(rle(turbidity_fnu), rep(lengths, lengths)),
    stuck  = rl_len >= STUCK_MIN_RUN
  ) %>%
  select(-rl_id, -rl_len)

#### Flag 2 — Spike: single-step rate of change exceeds threshold.
#### Retains genuine rapid storm rises (which are gradual across steps)
#### while catching instantaneous jumps caused by sensor fouling or
#### debris impacts.

raw_turb_flagged <- raw_turb_flagged %>%
  mutate(
    roc   = abs(turbidity_fnu - lag(turbidity_fnu)) / lag(turbidity_fnu),
    spike = !is.na(roc) & roc > SPIKE_ROC_THRESHOLD
  ) %>%
  select(-roc)

cat("Stuck-sensor flags :", sum(raw_turb_flagged$stuck,            na.rm = TRUE), "\n")
cat("Spike flags        :", sum(raw_turb_flagged$spike,            na.rm = TRUE), "\n")
cat("Total flagged      :", sum(raw_turb_flagged$stuck |
                                  raw_turb_flagged$spike, na.rm = TRUE), "\n")
cat("Total observations :", nrow(raw_turb_flagged), "\n")
```

```{r turb-qc-inspect, fig.width = 12, fig.height = 4}
#### Plot flagged observations over the full record to review before removing.
map(year(raw_turb_flagged$dateTime), 
    ~ raw_turb_flagged %>%
      filter(year(dateTime) == .x) %>% 
      mutate(flag = case_when(
        stuck & spike ~ "both",
        stuck         ~ "stuck",
        spike         ~ "spike",
        TRUE          ~ "ok"
      )) %>%
      ggplot(aes(x = dateTime, y = turbidity_fnu)) +
      geom_line(data = ~ filter(., flag == "ok"),
                linewidth = 0.2, alpha = 0.4, color = "steelblue") +
      geom_point(data = ~ filter(., flag != "ok"),
                 aes(color = flag), size = 1, alpha = 0.8) +
      scale_color_manual(values = c(stuck = "goldenrod3",
                                    spike = "firebrick",
                                    both  = "purple"),
                         name = "Flag") +
      #scale_y_log10() +
      labs(title = "Flagged turbidity observations",
           subtitle = paste0("stuck threshold: ", STUCK_MIN_RUN,
                             " consecutive identical values  |  ",
                             "spike threshold: ROC > ", SPIKE_ROC_THRESHOLD, "×"),
           x = NULL, y = "FNU") +
      theme_bw()
)
```

```{r turb-qc-clean}
#### Remove flagged observations.
#### Downstream, the flow-weighted aggregation step's na.approx() call will
#### interpolate across resulting gaps of ≤ 24 h (same as Kemper).

raw_turb_clean <- raw_turb_flagged %>%
  filter(!(stuck | spike)) %>%
  select(dateTime, turbidity_fnu)

cat("Removed :", nrow(raw_turb) - nrow(raw_turb_clean), "observations\n")
cat("Retained:", nrow(raw_turb_clean), "observations\n")
```

---

# Flow-Weighted Turbidity Aggregation (15-min → hourly)

```{r fw-aggregation}

################################################################################
#### This replicates the Kemper et al. approach:
####   1. Join instantaneous (15-min) turbidity to the hourly discharge record.
####      Each 15-min reading is assigned the flow of its containing hour.
####   2. Compute turbidity flux = turbidity × flow_cms.
####   3. Integrate flux and water volume over each hour using the time step
####      implied by observations per hour (1, 2, 3, or 4 per hour).
####   4. Flow-weighted mean = Σ(turbidity_load) / Σ(water_load) per hour.
####   5. Fill gaps ≤ 24 h by linear interpolation; log-transform.
################################################################################

#### Assign each 15-min turbidity reading the interpolated hourly flow value
#### by flooring the 15-min timestamp to the hour.

turb_with_flow <- raw_turb_clean %>%
  mutate(hour_dt = floor_date(dateTime, unit = "hour")) %>%
  inner_join(flow_hourly %>% 
               select(dateTime, flow_cms),
             by = c("hour_dt" = "dateTime")) %>%
  mutate(
    date = date(dateTime),
    hour = hour(dateTime)
  )

#### Count observations per hour (needed for correct time-step weighting)

turb_obs_per_hour <- turb_with_flow %>%
  mutate(obs_per_hour = n(),
         .by = c(date,hour)) 

#### Compute turbidity flux and water load for each 15-min observation

turb_flux <- turb_obs_per_hour %>%
  mutate(
    turbidity_flux = turbidity_fnu * flow_cms,
    turbidity_load = case_when(
      obs_per_hour == 4 ~ turbidity_flux * 60 * 15,
      obs_per_hour == 3 ~ turbidity_flux * 60 * 20,
      obs_per_hour == 2 ~ turbidity_flux * 60 * 30,
      obs_per_hour == 1 ~ turbidity_flux * 60 * 60
    ),
    water_load = case_when(
      obs_per_hour == 4 ~ flow_cms * 60 * 15,
      obs_per_hour == 3 ~ flow_cms * 60 * 20,
      obs_per_hour == 2 ~ flow_cms * 60 * 30,
      obs_per_hour == 1 ~ flow_cms * 60 * 60
    )
  )

#### Aggregate to hourly flow-weighted mean turbidity

hourly_turbidity <- turb_flux %>%
  summarise(
    dateTime = floor_date(dateTime[1], unit = "hour"),
    fw_mean_observed_turbidity = sum(turbidity_load) / sum(water_load),
    .by = c(date, hour)
  ) %>%
  ### Rebuild as a regular hourly tsibble and fill short gaps
  as_tsibble(index = "dateTime") %>%
  fill_gaps(.full = FALSE) %>%
  mutate(
    fw_mean_observed_turbidity = na.approx(fw_mean_observed_turbidity, 
                                           maxgap = 24, 
                                           na.rm = FALSE)
  ) %>%
  as_tibble() %>%
  mutate(
    log_observed_turbidity = log10(fw_mean_observed_turbidity)
  ) %>%
  filter(!is.na(fw_mean_observed_turbidity))

glimpse(hourly_turbidity)
```


# Merge Turbidity and Discharge

```{r merge}
#### Join flow-weighted hourly turbidity with the hourly interpolated flow.

hourly_merged <- hourly_turbidity %>%
  inner_join(., flow_hourly) %>%
  mutate(
    date = date(dateTime),
    hour = hour(dateTime),
    water_year = if_else(month(dateTime) >= 10,
                         year(dateTime) + 1L,
                         year(dateTime))
  ) %>%
  arrange(dateTime)

glimpse(hourly_merged)
```

---

# Calculate Antecedent Discharge Features

```{r antecedent-short}

################################################################################
#### Antecedent discharge features — short to medium term.
#### Window sizes are in hours, matching the Kemper et al. convention:
####
####   six_hour_lag_log_q         mean of prior 6 h
####   mean_prior_daily_log_q     mean of prior 24 h
####   mean_prior_weekly_log_q    mean of prior 168 h (7 days)
####   mean_prior_monthly_log_q   mean of prior 720 h (30 days)
####   max_prior_monthly_log_q    max  of prior 720 h
####   log_limb                   log(Q_t) − log(Q_{t-1})
####   dq_q                       log_limb / log(Q_t)
################################################################################

expanded_discharge <- hourly_merged %>%
  arrange(dateTime) %>%
  mutate(
    six_hour_lag_log_q = rollapply(log_observed_flow,
                                   width = list(-(6:1)), mean,
                                   align = "right", fill = NA),
    
    mean_prior_daily_log_q = rollapply(log_observed_flow,
                                       width = list(-(24:1)), mean,
                                       align = "right", fill = NA),
    
    mean_prior_weekly_log_q = rollapply(log_observed_flow,
                                        width = list(-((24*7):1)), mean,
                                        align = "right", fill = NA),
    
    mean_prior_monthly_log_q = rollapply(log_observed_flow,
                                         width = list(-((24*30):1)), mean,
                                         align = "right", fill = NA),
    
    max_prior_monthly_log_q = rollapply(log_observed_flow,
                                        width = list(-((24*30):1)), max,
                                        align = "right", fill = NA),
    
    log_limb = log_observed_flow - lag(log_observed_flow),
    dq_q = log_limb / log_observed_flow
  )
```

```{r antecedent-long}

################################################################################
#### Antecedent discharge features — long term.
#### These require larger rolling windows and take more time to compute.
################################################################################

expanded_discharge_long <- expanded_discharge %>%
  arrange(dateTime) %>%
  mutate(
    max_prior_threemonth_log_q = rollapply(log_observed_flow,
                                           width = list(-((24*30*3):1)), max,
                                           align = "right", fill = NA),
    
    max_prior_sixmonth_log_q = rollapply(log_observed_flow,
                                         width = list(-((24*30*6):1)), max,
                                         align = "right", fill = NA)
  ) %>%
  mutate(
    date_3month_max = slide2(
      log_observed_flow, dateTime,
      .f = ~ as_datetime(.y[which.max(.x)]),
      .before = (24*30*3), .complete = FALSE
    ) %>% map_dbl(as.numeric) %>% as_datetime(tz = "America/Denver"),
    
    date_6month_max = slide2(
      log_observed_flow, dateTime,
      .f = ~ as_datetime(.y[which.max(.x)]),
      .before = (24*30*6), .complete = FALSE
    ) %>% map_dbl(as.numeric) %>% as_datetime(tz = "America/Denver")
  )
```

```{r antecedent-annual}

################################################################################
#### Annual antecedent — this is the most time-consuming window.
################################################################################

annual_expanded_discharge <- expanded_discharge_long %>%
  mutate(
    max_prior_annual_q = rollapply(log_observed_flow,
                                   width = list(-((24*365):1)), max,
                                   align = "right", fill = NA)
  ) %>%
  mutate(
    date_year_max = slide2(
      log_observed_flow, dateTime,
      .f = ~ as_datetime(.y[which.max(.x)]),
      .before = (24*365), .complete = FALSE
    ) %>% map_dbl(as.numeric) %>% as_datetime(tz = "America/Denver")
  )
```

```{r peak-timing}

#### Time since the 3-month, 6-month, and annual discharge peaks (hours)

longer_term_expanded_discharge <- annual_expanded_discharge %>%
  mutate(
    time_since_3month_peak = abs(difftime(date_3month_max, dateTime, units = "hours")),
    time_since_6month_peak = abs(difftime(date_6month_max, dateTime, units = "hours")),
    time_since_annual_peak = abs(difftime(date_year_max, dateTime, units = "hours"))
  ) %>%
  select(
    -date_3month_max, -date_6month_max, -date_year_max
  )
```


# Finalize Modeling Dataset

```{r final-dataset}

################################################################################
#### Assemble the final modeling dataset, add DOY and water year, drop NA rows.
################################################################################

intake_turbidity_plus_drivers_clean <- longer_term_expanded_discharge %>%
  mutate(
    doy = yday(dateTime),
    month = month(dateTime),
    water_year = as.integer(water_year)
  ) %>%
  select(
    dateTime, date, hour, doy, month, water_year,
    fw_mean_observed_turbidity,
    log_observed_turbidity,
    flow_cms,
    log_observed_flow,
    log_squared_observed_flow,
    six_hour_lag_log_q,
    mean_prior_daily_log_q,
    mean_prior_weekly_log_q,
    mean_prior_monthly_log_q,
    max_prior_monthly_log_q,
    max_prior_threemonth_log_q,
    max_prior_sixmonth_log_q,
    max_prior_annual_q,
    log_limb, dq_q,
    time_since_3month_peak,
    time_since_6month_peak,
    time_since_annual_peak
  ) %>%
  drop_na()

cat("Final modeling dataset:", nrow(intake_turbidity_plus_drivers_clean),
    "rows,", ncol(intake_turbidity_plus_drivers_clean), "columns\n")
cat("Date range:",
    as.character(min(intake_turbidity_plus_drivers_clean$dateTime)), "to",
    as.character(max(intake_turbidity_plus_drivers_clean$dateTime)), "\n")
cat("Water years:",
    min(intake_turbidity_plus_drivers_clean$water_year), "to",
    max(intake_turbidity_plus_drivers_clean$water_year), "\n")

glimpse(intake_turbidity_plus_drivers_clean)
```


# Quick Diagnostic Plots

```{r plot-ts, fig.width=12, fig.height=4}

#### Hourly time series of flow-weighted turbidity

intake_turbidity_plus_drivers_clean %>%
  ggplot(aes(x = dateTime, y = fw_mean_observed_turbidity)) +
  geom_line(color = "steelblue", linewidth = 0.2, alpha = 0.6) +
  scale_y_log10() +
  labs(title = "FC Intake — Hourly Flow-Weighted Turbidity (log scale)",
       x = NULL, y = "Turbidity (FNU)") +
  theme_bw()
```

```{r plot-rating, fig.width=6, fig.height=5}

#### Rating curve: log(Q) vs. log(turbidity)

intake_turbidity_plus_drivers_clean %>%
  ggplot(aes(x = log_observed_flow, y = log_observed_turbidity)) +
  geom_point(alpha = 0.1, size = 0.5, color = "steelblue") +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2),
              color = "firebrick", se = TRUE) +
  labs(title = "Rating Curve — Canyon Mouth Q vs. FC Intake Turbidity",
       x = "log10(Q, m³/s)", y = "log10(Turbidity, FNU)") +
  theme_bw()
```


# Save data

```{r}
write_parquet(intake_turbidity_plus_drivers_clean, here(
  paste0("data/derived/models/turbidity_forecast/intake_turbidity_plus_drivers_clean_v", 
         Sys.Date(), ".parquet")
))
```


