---
title: "02_model_development — FC Intake Turbidity (Kemper Method)"
author: "ROSSyndicate"
date: "`r Sys.Date()`"
output: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

This script applies the Kemper et al. (2025) turbidity forecasting approach to
the Fort Collins Intake turbidity and Canyon Mouth discharge data prepared in
`01_data_prep.Rmd`. Data are at **hourly** resolution (flow-weighted turbidity
aggregated from 15-min SCADA data, hourly discharge from CDSSR telemetry).

Two models are developed (analogous to Kemper's Quantile-Coldbrook and
LightGBM-Coldbrook models):

1. **Quantile-FC** — Median quantile regression of log(turbidity) on
log(Q) + log²(Q), following Mukundan et al. (2018) / Wang et al. (2021).
Simple, interpretable, single-predictor model.

2. **LightGBM-FC** — Gradient-boosted decision tree using all antecedent
discharge predictors. Uses backwards variable selection and expanding-
window cross-validation for hyperparameter tuning.

Because only one flow station is available, there is no network-model
equivalent (LightGBM-Network).

Standard k-fold cross-validation is inappropriate here due to temporal autocorrelation and
explicit lag predictors (e.g., antecedent discharge features) — randomly shuffling folds
would allow training data to be temporally "after" validation data, leaking future information.
Instead, LightGBM variable selection and hyperparameter tuning use expanding-window
(forward-chaining) cross-validation, which respects temporal ordering:

Variable selection (expanding window, 2 folds):
- WY2012–WY2020 train → WY2021 validate
- WY2012–WY2021 train → WY2022 validate

Hyperparameter tuning (expanding window, 3 folds):
- WY2012–WY2019 train → WY2020 validate
- WY2012–WY2020 train → WY2021 validate
- WY2012–WY2021 train → WY2022 validate

The final models are retrained on WY2011–WY2022 (training + validation) before
evaluation on the held-out test period (WY2023–WY2025).

################################################################################


```{r packages}
library(here)
library(tidyverse)
library(arrow)
library(crayon)

### Modeling
library(quantreg)
library(hydroGOF)
library(lightgbm)

### Tuning (tidymodels stack)
library(tidymodels)
library(bonsai)
```

### Load prepared data

```{r load-data}
intake_turbidity_plus_drivers_clean <- read_parquet(here("data/derived/models/turbidity_forecast",
                                                         "intake_turbidity_plus_drivers_clean_v2026-03-25.parquet"))
```


# Helper functions

```{r helpers}

#### Decompose KGE into correlation, bias, and variability components
#### (Gupta et al. 2009, https://doi.org/10.1016/j.jhydrol.2009.08.003)

decompose_kge <- function(sim, obs) {
  r <- cor(sim, obs, method = "pearson", use = "pairwise.complete.obs")
  bias <- mean(sim, na.rm = TRUE) / mean(obs, na.rm = TRUE)
  variability <- sd(sim, na.rm = TRUE)   / sd(obs, na.rm = TRUE)
  
  kge_manual <- 1 - sqrt((r - 1)^2 + (variability - 1)^2 + (bias - 1)^2)
  
  tibble(r, variability, bias, kge_manual)
}


#### Compute a standard suite of model performance metrics

error_estimator <- function(pred, obs, model_name) {
  raw_err <- pred - obs
  abs_err <- abs(raw_err)
  sqr_err <- raw_err^2
  pct_err <- abs_err / obs
  
  tibble(
    model = model_name,
    rmse = sqrt(mean(sqr_err, na.rm = TRUE)),
    mae = mean(abs_err,  na.rm = TRUE),
    mape = mean(pct_err,  na.rm = TRUE),
    nse = NSE(pred, obs),
    kge = KGE(pred, obs),
    pbias = pbias(pred, obs),
    corr = cor(pred, obs, use = "complete.obs")
  ) %>%
    bind_cols(decompose_kge(pred, obs))
}
```


# Data Splits

```{r splits}

#### Training data (WY2011–WY2021)
intake_train <- intake_turbidity_plus_drivers_clean %>%
  filter(water_year >= 2011, water_year <= 2021)

#### Validation data (WY2022)
intake_valid <- intake_turbidity_plus_drivers_clean %>%
  filter(water_year == 2022)

#### Test data (WY2023–WY2025)
intake_test  <- intake_turbidity_plus_drivers_clean %>%
  filter(water_year >= 2023)

#### Training + validation combined (for final model training before test eval)
intake_train_full <- intake_turbidity_plus_drivers_clean %>%
  filter(water_year <= 2022)

cat("Train rows:", nrow(intake_train),
    "| Valid rows:", nrow(intake_valid),
    "| Test rows:", nrow(intake_test), "\n")
```

---

# Model 1: Quantile Regression (Quantile-FC)

```{r qqm-build}

################################################################################
#### Median (tau = 0.5) quantile regression of log10(turbidity) on
#### log10(Q) and log10²(Q).
####
#### Model form (Mukundan et al. 2018 / Wang et al. 2021):
####   log10(T) = β0 + β1·log10(Q) + β2·log10²(Q)
####
#### Trained on the full training+validation dataset (WY2011–WY2022)
#### so that the test evaluation uses data the model has never seen.
################################################################################

#### Set up the tau we care about (p75)

qqm_taus <- tibble(tau = 0.75) %>%
  bind_cols(.,
            intake_train_full %>%
              select(log_observed_turbidity,
                     log_observed_flow,
                     log_squared_observed_flow) %>%
              filter(if_all(everything(), ~ is.finite(.x))) %>%
              nest()
  )

#### Fit quantile regression at tau = 0.75

qqm_raw <- qqm_taus %>%
  mutate(model = map2(data, tau,
                      ~ rq(log_observed_turbidity ~ 1 +
                             log_observed_flow + log_squared_observed_flow,
                           data = .x,
                           tau  = .y)))

#### Extract coefficients

qqm_coefficients <- qqm_raw %>%
  mutate(
    beta0 = map_dbl(model, ~ .x$coefficients[[1]]),
    beta1 = map_dbl(model, ~ .x$coefficients[[2]]),
    beta2 = map_dbl(model, ~ .x$coefficients[[3]])
  ) %>%
  select(tau, beta0, beta1, beta2)

print(qqm_coefficients)
```

```{r qqm-smearing}

#### Predict on training+validation data to estimate the Duan smearing factor.
#### The smearing factor corrects for the systematic underestimation that arises
#### when back-transforming from log scale (Duan 1983).

qqm_train_pred <- qqm_raw %>%
  mutate(data = map2(data, model,
                     ~ mutate(.x,
                              log_pred_turb = .y$coefficients[[1]] +
                                .y$coefficients[[2]] * log_observed_flow +
                                .y$coefficients[[3]] * log_squared_observed_flow))) %>%
  unnest(data)

D_qqm <- qqm_train_pred %>%
  mutate(log_resids = 10^(log_pred_turb - log_observed_turbidity)) %>%
  summarise(D = mean(log_resids, na.rm = TRUE)) %>%
  pull(D)

cat("Duan smearing factor (D_qqm):", round(D_qqm, 4), "\n")
```

```{r qqm-test-predict}

#### Predict on test data

qqm_test_pred <- intake_test %>%
  mutate(
    log_pred_turb  = qqm_coefficients$beta0 +
      qqm_coefficients$beta1 * log_observed_flow +
      qqm_coefficients$beta2 * log_squared_observed_flow,
    pred_turb_fnu  = (10^log_pred_turb) * D_qqm
  )
```

```{r qqm-evaluate}

#### Evaluate quantile model on test data

qqm_stats <- error_estimator(
  pred       = qqm_test_pred$pred_turb_fnu,
  obs        = qqm_test_pred$fw_mean_observed_turbidity,
  model_name = "Quantile-FC"
)

print(qqm_stats)
```

```{r qqm-plot, fig.width=10, fig.height=4}

#### Time series plot of test-period predictions vs. observations
#### Aggregated to daily: observed p75 vs. predicted mean (tau = 0.75 model)

qqm_test_pred %>%
  mutate(date = as.Date(dateTime)) %>%
  group_by(date) %>%
  summarise(
    obs_p75  = quantile(fw_mean_observed_turbidity, 0.75, na.rm = TRUE),
    pred_mean = mean(pred_turb_fnu, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(obs_p75, pred_mean),
               names_to = "series", values_to = "turbidity") %>%
  mutate(series = recode(series,
                         obs_p75   = "Observed (daily p75)",
                         pred_mean = "Quantile-FC (daily mean)")) %>%
  ggplot(aes(x = date, y = turbidity, color = series)) +
  geom_line(linewidth = 0.5, alpha = 0.8) +
  scale_color_manual(values = c("Observed (daily p75)"      = "black",
                                "Quantile-FC (daily mean)"  = "firebrick")) +
  labs(title = "Quantile-FC: Test Period Predictions vs. Observations (daily)",
       x = NULL, y = "Turbidity (FNU)", color = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```


# Model 2: LightGBM-FC

## Variable Selection

```{r lgbm-predictor-setup}

################################################################################
#### Build the full predictor matrix for backwards variable selection.
#### We include all antecedent discharge features plus day-of-year, month,
#### and water_year (as an integer to capture long-term trends).
#### The response is log10(turbidity).
################################################################################

#### All candidate predictors (drop raw/non-predictor columns)

lgbm_predictors <- intake_turbidity_plus_drivers_clean %>%
  select(
    log_observed_turbidity,
    water_year,
    hour, doy, month,
    log_observed_flow,
    log_squared_observed_flow,
    six_hour_lag_log_q,
    mean_prior_daily_log_q,
    mean_prior_weekly_log_q,
    mean_prior_monthly_log_q,
    max_prior_monthly_log_q,
    max_prior_threemonth_log_q,
    max_prior_sixmonth_log_q,
    max_prior_annual_q,
    log_limb, dq_q,
    time_since_3month_peak,
    time_since_6month_peak,
    time_since_annual_peak
  ) %>%
  mutate(water_year2 = water_year) %>%  # allow water_year to be a predictor
  filter(is.finite(log_observed_turbidity))  # drop zero-turbidity rows (log10(0) = -Inf)
# while keeping it for train/test splits
```

```{r lgbm-varsel, eval=FALSE}

################################################################################
#### Backwards variable selection.
#### Iteratively train LightGBM with default hyperparameters,
#### calculate feature importance (mean Gain), remove the least important
#### feature, and repeat. Records NSE, KGE, and PBIAS for each iteration
#### across an expanding validation window (WY2019→2021, WY2020→2022).
################################################################################

model_stats_lgbm <- list()
lgbm_var_imp <- list()
vars_list <- list()
all_model_stats <- list()

#### Default LightGBM hyperparameters for variable selection phase

lgbm_default_params <- list(
  objective = "regression",
  num_leaves = 31L,
  learning_rate = 0.1,
  min_data_in_leaf = 20L,
  num_threads = 4L,
  verbose = -1L
)


for (i in seq_len(ncol(lgbm_predictors) - 2L)) {  # -2 for response + water_year2
  
  run <- paste0("run", i)
  cat(cyan("\nVariable Selection Run", i,
           "| Predictors remaining:",
           ncol(lgbm_predictors) - 2L, "\n"))
  
  for (j in 2021:2022) {
    
    cat(yellow("  Validating on WY", j, "\n"))
    
    train_yrs <- seq(2012, j - 1L, 1L)
    
    predictors_train <- lgbm_predictors %>%
      filter(water_year %in% train_yrs)
    
    predictors_test <- lgbm_predictors %>%
      filter(water_year == j) %>%
      select(colnames(lgbm_predictors))
    
    #### Predictor matrix (drop response + water_year control columns)
    preds_mat <- data.matrix(predictors_train %>%
                               select(!c(log_observed_turbidity,
                                         water_year,
                                         water_year2)))
    response_vec <- predictors_train$log_observed_turbidity
    
    lgbm_train_ds <- lgb.Dataset(preds_mat, label = response_vec)
    
    test_mat <- data.matrix(predictors_test %>%
                              select(!c(log_observed_turbidity,
                                        water_year,
                                        water_year2)))
    
    #### Train
    set.seed(913)
    lgbm_mod <- lgb.train(lgbm_default_params,
                          data = lgbm_train_ds,
                          nrounds = 100L,
                          verbose = -1L)
    
    #### Predict + smear
    log_pred_train <- predict(lgbm_mod, newdata = preds_mat)
    smear_d <- mean(10^(response_vec - log_pred_train), na.rm = TRUE)
    log_pred <- predict(lgbm_mod, newdata = test_mat)
    pred_turb <- (10^log_pred) * smear_d
    obs_turb <- 10^predictors_test$log_observed_turbidity
    
    #### Performance
    stats_row <- error_estimator(pred_turb, obs_turb,
                                 paste0(run, "_WY", j))
    
    model_stats_lgbm[[paste0(run, "_WY", j)]] <- stats_row
    
    #### Feature importance (mean Gain)
    imp <- lgb.importance(lgbm_mod, percentage = TRUE) %>%
      as_tibble() %>%
      mutate(run = run, valid_wy = j)
    
    lgbm_var_imp[[paste0(run, "_WY", j)]] <- imp
    
  }  # end year loop
  
  #### Identify and drop least important feature
  mean_imp <- bind_rows(lgbm_var_imp[grep(run, names(lgbm_var_imp))]) %>%
    group_by(Feature) %>%
    summarise(mean_gain = mean(Gain, na.rm = TRUE)) %>%
    arrange(mean_gain)
  
  vars_list[[run]] <- lgbm_predictors %>%
    select(!c(log_observed_turbidity, water_year, water_year2)) %>%
    colnames()
  
  all_model_stats[[run]] <- bind_rows(model_stats_lgbm[grep(run,
                                                            names(model_stats_lgbm))])
  
  #### Drop the least important feature for the next iteration
  drop_var <- mean_imp$Feature[[1]]
  cat(red("    Dropping:", drop_var, "\n"))
  
  lgbm_predictors <- lgbm_predictors %>%
    select(-all_of(drop_var))
  
  #### Stop if only one predictor remains
  if ((ncol(lgbm_predictors) - 2L) <= 1L) break
  
}  # end variable-selection loop

all_varsel_stats <- bind_rows(all_model_stats, .id = "run")
```

```{r lgbm-select-model, eval=FALSE}

################################################################################
#### Identify the best model from variable selection.
#### Selection criterion (following Kemper et al.):
####   - NSE > 0 (better than mean model)
####   - KGE > 0
####   - |PBIAS| < 25%
#### Among models meeting all three criteria, choose the one with the fewest
#### predictors (parsimony) and the highest mean KGE.
################################################################################

#### Summarise performance across validation years for each run

varsel_summary <- all_varsel_stats %>%
  group_by(run) %>%
  summarise(
    mean_nse = mean(nse, na.rm = TRUE),
    mean_kge = mean(kge, na.rm = TRUE),
    mean_pbias = mean(pbias, na.rm = TRUE)
  ) %>%
  filter(mean_nse > 0, mean_kge > 0, abs(mean_pbias) < 25) %>%
  arrange(desc(mean_kge))

best_run <- varsel_summary$run[[1]]
cat("Selected model run:", best_run, "\n")

#### Retrieve the predictor list for the best model

best_predictors <- vars_list[[best_run]]
cat("Predictors in best model:", paste(best_predictors, collapse = ", "), "\n")
```

## Hyperparameter Tuning

```{r lgbm-tune, eval=FALSE}

################################################################################
#### Tune LightGBM hyperparameters using an expanding-window design via
#### tidymodels. Parameters tuned: min_n, num_leaves.
#### Grid: 100-point max-entropy design over the search space.
################################################################################

#### Prepare data with only the selected predictors

lgbm_tune_data <- intake_turbidity_plus_drivers_clean %>%
  select(all_of(c("log_observed_turbidity", "water_year",
                  best_predictors))) %>%
  mutate(water_year2 = water_year)

#### Build expanding-window train/validation splits
#### WY2012–WY2019 train → WY2020 valid
#### WY2012–WY2020 train → WY2021 valid
#### WY2012–WY2021 train → WY2022 valid

splits_list <- list()

for (h in 2020:2022) {
  train_d <- lgbm_tune_data %>%
    filter(water_year >= 2012, water_year < h) %>%
    select(!c(water_year, water_year2))
  
  valid_d <- lgbm_tune_data %>%
    filter(water_year == h) %>%
    select(!c(water_year, water_year2))
  
  k <- 2023 - h
  splits_list[[k]] <- make_splits(train_d, assessment = valid_d)
}

tune_resamples <- manual_rset(splits_list, c("Split1", "Split2", "Split3"))

#### Model recipe and spec

tune_recipe <- recipe(
  log_observed_turbidity ~ .,
  data = lgbm_tune_data %>%
    filter(water_year >= 2012, water_year < 2020) %>%
    select(!c(water_year, water_year2))
) %>%
  prep()

lgbm_spec <- boost_tree(
  mode = "regression",
  min_n = tune(),
) %>%
  set_engine("lightgbm",
             objective = "rmse",
             verbose = -1,
             num_leaves = tune(),
             num_threads = 4L)

#### Parameter grid

lgbm_params <- parameters(
  min_n(c(2L, 60L), trans = NULL),
  num_leaves(c(2L, 256L))
)

lgbm_grid <- grid_max_entropy(lgbm_params, size = 100)

#### Workflow

lgbm_wf <- workflow() %>%
  add_model(lgbm_spec) %>%
  add_formula(log_observed_turbidity ~ .)

#### Tune

set.seed(913)
lgbm_tuned <- tune_grid(
  object = lgbm_wf,
  resamples = tune_resamples,
  grid = lgbm_grid,
  metrics = metric_set(rmse),
  control = control_grid(verbose = TRUE)
)

#### Extract best hyperparameters

rmse_scores <- lgbm_tuned %>%
  show_best(metric = "rmse", n = 100) %>%
  mutate(
    mean_rmse_rank = dense_rank(mean),
    se_rank = dense_rank(std_err),
    combined_rank = mean_rmse_rank + se_rank
  ) %>%
  arrange(combined_rank)

best_params <- rmse_scores %>% slice(1)
cat("Best min_n:", best_params$min_n,
    "| Best num_leaves:", best_params$num_leaves, "\n")
```

## Train Final LightGBM Model

```{r lgbm-final-train, eval=FALSE}

################################################################################
#### Train the final LightGBM model on all data up to WY2022 using the
#### selected predictors and tuned hyperparameters, then evaluate on WY2023+.
################################################################################

#### Prepare training and test matrices

lgbm_final_train <- intake_train_full %>%
  filter(water_year >= 2011) %>%
  select(all_of(c(best_predictors, "log_observed_turbidity")))

lgbm_final_test <- intake_test %>%
  select(all_of(c(best_predictors, "log_observed_turbidity")))

train_mat <- data.matrix(lgbm_final_train %>%
                           select(!log_observed_turbidity))
train_label <- lgbm_final_train$log_observed_turbidity

test_mat_final <- data.matrix(lgbm_final_test %>%
                                select(!log_observed_turbidity))

lgbm_ds <- lgb.Dataset(train_mat, label = train_label)

#### Final hyperparameters (replace with tuned values after running tune chunk)

final_params <- list(
  objective = "regression",
  num_leaves = best_params$num_leaves,   # from tuning
  min_data_in_leaf = best_params$min_n,        # from tuning
  learning_rate = 0.1,
  num_threads = 4L,
  verbose = -1L
)

set.seed(913)
lgbm_final_model <- lgb.train(final_params,
                              data = lgbm_ds,
                              nrounds = 100L)

#### Smearing factor from training data

log_pred_train <- predict(lgbm_final_model, data = train_mat)
D_lgbm <- mean(10^(train_label - log_pred_train), na.rm = TRUE)
cat("Duan smearing factor (D_lgbm):", round(D_lgbm, 4), "\n")

#### Predict on test data

log_pred_test <- predict(lgbm_final_model, data = test_mat_final)
pred_turb_lgbm <- (10^log_pred_test) * D_lgbm
obs_turb_test <- 10^lgbm_final_test$log_observed_turbidity

#### Save the model

lgb.save(lgbm_final_model,
         filename = here("modeling/turbidity/kemper_application/lgbm_fc_model.txt"))
```

## Evaluate LightGBM Model

```{r lgbm-evaluate, eval=FALSE}

lgbm_stats <- error_estimator(pred_turb_lgbm, obs_turb_test, "LightGBM-FC")
print(lgbm_stats)

```

```{r lgbm-variable-importance, eval=FALSE}

#### Variable importance plot

lgb.importance(lgbm_final_model, percentage = TRUE) %>%
  as_tibble() %>%
  slice_max(Gain, n = 15) %>%
  ggplot(aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "LightGBM-FC: Top 15 Features by Mean Gain",
       x = NULL, y = "Mean Gain (%)") +
  theme_bw()
```

```{r lgbm-plot, eval=FALSE, fig.width=10, fig.height=4}

#### Test period time series comparison

tibble(
  dateTime  = intake_test$dateTime,
  observed  = obs_turb_test,
  predicted = pred_turb_lgbm
) %>%
  pivot_longer(cols = c(observed, predicted),
               names_to = "series", values_to = "turbidity") %>%
  ggplot(aes(x = dateTime, y = turbidity, color = series)) +
  geom_line(linewidth = 0.5, alpha = 0.8) +
  scale_y_log10() +
  scale_color_manual(values = c("observed" = "black",
                                "predicted" = "steelblue")) +
  labs(title = "LightGBM-FC: Test Period Predictions vs. Observations",
       x = NULL, y = "Turbidity (FNU, log scale)", color = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

---

# Model Comparison

```{r compare, eval=FALSE}

################################################################################
#### Side-by-side performance summary for both models on the test period.
################################################################################

model_comparison <- bind_rows(qqm_stats, lgbm_stats) %>%
  select(model, rmse, mae, nse, kge, pbias, corr, r, variability, bias)

print(model_comparison)
```


```{r scatter-compare, eval=FALSE, fig.width=8, fig.height=4}
#### Observed vs. predicted scatter for both models

qqm_scatter <- tibble(
  model = "Quantile-FC",
  observed = intake_test$fw_mean_observed_turbidity,
  predicted = qqm_test_pred$pred_turb_fnu
)

lgbm_scatter <- tibble(
  model = "LightGBM-FC",
  observed = obs_turb_test,
  predicted = pred_turb_lgbm
)

bind_rows(qqm_scatter, lgbm_scatter) %>%
  ggplot(aes(x = observed, y = predicted, color = model)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_manual(values = c("Quantile-FC" = "firebrick",
                                "LightGBM-FC" = "steelblue")) +
  facet_wrap(~ model) +
  labs(title = "Observed vs. Predicted Turbidity — Test Period (WY2023–2025)",
       x = "Observed (FNU)", y = "Predicted (FNU)") +
  theme_bw() +
  theme(legend.position = "none")
```
